### 2024-11-26 paper/paper.qmd
Upload starter files

---
title: "financial Well Being Analysis"
# subtitle: "My subtitle if needed"
author: 
  - Xinxiang Gao
thanks: "Code and data are available at: [https://github.com/xgao28/financial_well_being_analysis](https://github.com/xgao28/financial_well_being_analysis)."
date: today
date-format: long
abstract: "This should be an abstract. The report is still being finalized."
format: pdf
number-sections: true
bibliography: references.bib
---

```{r}
#| include: false
#| warning: false
#| message: false

library(tidyverse)
library(arrow)
data <- read_parquet("../data/02-analysis_data/NFWBS_analysis_data.parquet")
```

# Introduction

# Data

```{r}

```

# Model

# Results

# Discussion

# Appendix

## Data cleaning

## Surveys, sampling, and observational data

# References

### 2024-11-26 scripts/02-clean_data.R

@@ -1,0 +1,1401 @@
+# The plyr library will be used to help us remap values
+library(plyr)
+
+# Load the data in to a dataframe for later use, please enter the appropriate path for the CSV.
+df <- read.csv('../data/01-raw_data/NFWBS_PUF_2016_data.csv')
+
+# This code maps values from the dummy variables to the descriptors from the codebook.
+df$sample = revalue(factor(df$sample), c(
+  `1` = "General population",
+  `2` = "Age 62+ oversample",
+  `3` = "Race/ethnicity and poverty oversample"
+))
+df$fpl = revalue(factor(df$fpl), c(
+  `1` = "<100% FPL",
+  `2` = "100%-199% FPL",
+  `3` = "200%+ FPL"
+))
+df$SWB_1 = revalue(factor(df$SWB_1), c(
+  `-4` = "Response not written to database",
+  `-1` = "Refused",
+  `1` = "1 Strongly disagree",
+  `2` = "2",
+  `3` = "3",
+  `4` = "4",
+  `5` = "5",
+  `6` = "6",
+  `7` = "7 Strongly agree"
+))
+df$SWB_2 = revalue(factor(df$SWB_2), c(
+  `-4` = "Response not written to database",
+  `-1` = "Refused",
+  `1` = "1 Strongly disagree",
+  `2` = "2",
+  `3` = "3",
+  `4` = "4",
+  `5` = "5",
+  `6` = "6",
+  `7` = "7 Strongly agree"
+))
+df$SWB_3 = revalue(factor(df$SWB_3), c(
+  `-4` = "Response not written to database",
+  `-1` = "Refused",
+  `1` = "1 Strongly disagree",
+  `2` = "2",
+  `3` = "3",
+  `4` = "4",
+  `5` = "5",
+  `6` = "6",
+  `7` = "7 Strongly agree"
+))
+df$FWBscore = revalue(factor(df$FWBscore), c(
+  `-4` = "Response not written to database",
+  `-1` = "Refused"
+))
+df$FWB1_1 = revalue(factor(df$FWB1_1), c(
+  `-4` = "Response not written to database",
+  `-1` = "Refused",
+  `1` = "Not at all",
+  `2` = "Very little",
[…]

### 2024-11-26 paper/paper.qmd
Updated readme, paper on acknowledgements

@@
 library(arrow)
+library(ggplot2)
 data <- read_parquet("../data/02-analysis_data/NFWBS_analysis_data.parquet")
@@
 ```{r}
+#| echo: false
+#| warning: false
+#| message: false
+library(skimr)
+skim_summary <- data %>%
+  select(-PUF_ID) %>%  # Remove 'PUF_ID' column
+  select(where(is.numeric)) %>%  # Keep only numeric columns
+  skim()
 
+# Round the numeric columns of the skim summary
+skim_summary %>%
+  mutate_if(is.numeric, ~round(., 3)) %>%
+  select(-complete_rate, -n_missing, -numeric.hist) %>% 
+  print()
+```
+```{r}
+#| echo: false
+#| warning: false
+#| message: false
+data_cleaned <- data[, !names(data) %in% "puf_id"]
+```
+
+```{r}
+#| echo: false
+#| warning: false
+#| message: false
+data_long <- data_cleaned %>%
+  pivot_longer(
+    cols = c(FWBscore, FSscore, LMscore, KHscore),
+    names_to = "ScoreType", values_to = "Value"
+  )
+
+# Plot using ggplot2
+ggplot(data_long, aes(x = Value)) +
+  geom_histogram(binwidth = 1, color = "black", fill = "skyblue", alpha = 0.7) +
+  facet_wrap(~ScoreType, scales = "free", ncol = 2) +
+  labs(
+    title = "Histograms of Financial Scores",
+    caption = "FWBscore: Financial Well-being Scale\nFSscore: Financial Skill Scale\nLMscore: Lusardi and Mitchell Financial Knowledge Scale\nKHscore: Knoll and Houts Financial Knowledge Scale",
+    x = "Score Value",
+    y = "Frequency"
+  ) +
+  theme_minimal()
+```
+
+The composition of finalwt remains on investigation.
+
+```{r}
+#| echo: false
+#| warning: false
+#| message: false
+ggplot(data_cleaned, aes(x = finalwt)) +
+  geom_density(color = "black", fill = "skyblue", alpha = 0.7) +
+  labs(
+    title = "Distribution of Final Weight (finalwt)",
+    x = "Final Weight (finalwt)",
+    y = "Frequency"
+  ) +
+  theme_minimal()
 ```
@@
 # Model
+
+# Linear Regression Model
+
+The linear regression model can be expressed mathematically as:
+
+$$y_i = \beta_0 + \beta_1 \cdot \text{LMscore}_i + \beta_2 \cdot \text{KHscore}_i + \beta_3 \cdot \text{FWBscore}_i + \beta_4 \cdot \text{FSscore}_i + \epsilon_i, \quad \epsilon_i \sim \mathcal{N}(0, \sigma^{2})$$
+
+### Model Components
+- $y_i$: The dependent variable, $\text{finalwt}_i$, for observation $i$.
+- $\beta_0$: The intercept term (constant in the model).
+- $\beta_1, \beta_2, \beta_3, \beta_4$: The slopes (coefficients) of the respective predictor variables:
+  - $\text{LMscore}_i$: Lusardi and Mitchell financial knowledge scale score.
+  - $\text{KHscore}_i$: Knoll and Houts financial knowledge scale score.
+  - $\text{FWBscore}_i$: Financial well-being scale score.
+  - $\text{FSscore}_i$: Financial skill scale score.
+
+```{r}
+#| echo: false
+#| warning: false
+#| message: false
+library(knitr)
+# Build a linear regression model
+# Assuming other columns in the dataset are predictors
+model <- lm(finalwt ~ LMscore + KHscore + FWBscore + FSscore, data = data_cleaned)
+
+# Display the summary of the model
+model_summary <- summary(model)
+
+# Create a data frame with key statistics
+lm_results <- data.frame(
+  Term = rownames(model_summary$coefficients),
+  Estimate = model_summary$coefficients[, "Estimate"],
+  Std_Error = model_summary$coefficients[, "Std. Error"],
+  t_value = model_summary$coefficients[, "t value"],
+  p_value = model_summary$coefficients[, "Pr(>|t|)"]
+)
+
+# Round values for better readability
+lm_results[, -1] <- round(lm_results[, -1], 3)
+
+# Display the table using knitr::kable
+kable(
+  lm_results,
+  caption = "Summary of Linear Regression Model"
+)
+```
 
@@
 
+## Acknowledgements
+
+We would like to express our gratitude to the developers and contributors of R [@citeR] as well as several R packages that were essential for the analysis and visualization of the data in this report.
+The following R packages provided indispensable tools and functionalities:
+
+- `tidyverse` [@tidyverse]: A collection of R packages designed for data science, including dplyr, ggplot2, readr, purrr, and others, which greatly facilitated data manipulation, analysis, and visualization.
+
+- `ggplot2` [@ggplot2]: An implementation of the Grammar of Graphics, which allowed us to create complex and aesthetically pleasing visualizations with ease.
+
+- `knitr` [@knitr]: This package enabled us to perform data demonstration with tables.
+
+- `styler` [@styler]: This package is helpful for styling the code.
+
+- `arrow` [@arrow]: This package provides a convenient and efficient way to work with parquet format.
+
+We are grateful to the Consumer Financial Protection Bureau (CFPB) for conducting the Financial Well-Being Survey [@cfpb_fwb_survey] and making the data publicly available.
+Their dedication to collecting and sharing such vital information significantly contributes to research on financial well-being.
+
+Finally, we would like to thank all those who contributed to the development and maintenance of the R programming language and its ecosystem, as well as the broader open-source community, whose efforts make such research possible.
+
 # References

### 2024-11-26 scripts/02-clean_data.R

@@ -1,21 +1,30 @@
+# Purpose: Cleans the raw data after download
+# Author: CFPB, Xinxiang Gao
+# Date: 26 November 2024
+# Contact: xinxiang.gao@mail.utoronto.ca
+# License: MIT
+# Acknowledgements: some of the code below utilized provided Code for analytics tools from CFPB.
+# https://www.consumerfinance.gov/data-research/financial-well-being-survey-data/
+
+
 # The plyr library will be used to help us remap values
-library(plyr)
+library(dplyr)
 
 # Load the data in to a dataframe for later use, please enter the appropriate path for the CSV.
-df <- read.csv('../data/01-raw_data/NFWBS_PUF_2016_data.csv')
+df <- read.csv("data/01-raw_data/NFWBS_PUF_2016_data.csv")
 
 # This code maps values from the dummy variables to the descriptors from the codebook.
-df$sample = revalue(factor(df$sample), c(
+df$sample <- revalue(factor(df$sample), c(
   `1` = "General population",
   `2` = "Age 62+ oversample",
   `3` = "Race/ethnicity and poverty oversample"
 ))
-df$fpl = revalue(factor(df$fpl), c(
+df$fpl <- revalue(factor(df$fpl), c(
   `1` = "<100% FPL",
   `2` = "100%-199% FPL",
   `3` = "200%+ FPL"
 ))
-df$SWB_1 = revalue(factor(df$SWB_1), c(
+df$SWB_1 <- revalue(factor(df$SWB_1), c(
   `-4` = "Response not written to database",
   `-1` = "Refused",
   `1` = "1 Strongly disagree",
@@ -26,7 +35,7 @@ df$SWB_1 = revalue(factor(df$SWB_1), c(
   `6` = "6",
   `7` = "7 Strongly agree"
 ))
-df$SWB_2 = revalue(factor(df$SWB_2), c(
+df$SWB_2 <- revalue(factor(df$SWB_2), c(
   `-4` = "Response not written to database",
   `-1` = "Refused",
   `1` = "1 Strongly disagree",
[…]

### 2024-11-26 paper/paper.qmd
Updated tests and simulations

@@
 
+## Significance
+
+## Estimand
+
 # Data
+
+## Overview
+
+## Measurement
 
@@
 skim_summary <- data %>%
-  select(-PUF_ID) %>%  # Remove 'PUF_ID' column
+  #select(-PUF_ID) %>%  # Remove 'PUF_ID' column
   select(where(is.numeric)) %>%  # Keep only numeric columns
@@
 
-# Linear Regression Model
-
 The linear regression model can be expressed mathematically as:
@@
   - $\text{FSscore}_i$: Financial skill scale score.
+
+# Results
 
@@
 
-# Results
-
 # Discussion
@@
 
+- `validate` [@validate]: This package provide useful functions for data tests.
+
 We are grateful to the Consumer Financial Protection Bureau (CFPB) for conducting the Financial Well-Being Survey [@cfpb_fwb_survey] and making the data publicly available.

### 2024-11-26 scripts/00-simulate_data.R

@@ -1,0 +1,42 @@
+# Purpose: Simulate an expected dataset
+# Author: Xinxiang Gao
+# Date: 26 November 2024
+# Contact: xinxiang.gao@mail.utoronto.ca
+# License: MIT
+
+
+set.seed(2690)  # Set seed for reproducibility
+
+# Number of observations
+n <- 1000
+
+# Simulate FWBscore (random values between 1 and 79)
+FWBscore <- sample(1:79, n, replace = TRUE)
+
+# Simulate FSscore (random values between 1 and 77)
+FSscore <- sample(1:77, n, replace = TRUE)
+
+# Simulate LMscore (random values between 0 and 3)
+LMscore <- sample(0:3, n, replace = TRUE)
+
+# Simulate KHscore (random values between -2 and 1.267)
+KHscore <- runif(n, min = -2, max = 1.267)  # Uniform random distribution
+
+# Simulate finalwt (random values between 0.17 and 6.64)
+finalwt <- runif(n, min = 0, max = 6.64)  # Uniform random distribution
+
+# Simulate PUF_ID (random values between 1 and 14400)
+PUF_ID <- sample(7123:14400, n, replace = FALSE)
+
+# Create the data frame
+data_simulated <- data.frame(
+  PUF_ID = PUF_ID,
+  FWBscore = FWBscore,
+  FSscore = FSscore,
+  LMscore = LMscore,
+  KHscore = KHscore,
+  finalwt = finalwt
+)
+
+write.csv(data_simulated, "data/00-simulated_data/simulated_data.csv", row.names = FALSE)
+

### 2024-11-26 scripts/01-clean_data.R

@@ -1,0 +1,1417 @@
+# Purpose: Cleans the raw data after download
+# Author: CFPB, Xinxiang Gao
+# Date: 26 November 2024
+# Contact: xinxiang.gao@mail.utoronto.ca
+# License: MIT
+# Acknowledgements: some of the code below utilized provided Code for analytics tools from CFPB.
+# https://www.consumerfinance.gov/data-research/financial-well-being-survey-data/
+
+
+# The plyr library will be used to help us remap values
+library(dplyr)
+
+# Load the data in to a dataframe for later use, please enter the appropriate path for the CSV.
+df <- read.csv("data/01-raw_data/NFWBS_PUF_2016_data.csv")
+
+# This code maps values from the dummy variables to the descriptors from the codebook.
+df$sample <- revalue(factor(df$sample), c(
+  `1` = "General population",
+  `2` = "Age 62+ oversample",
+  `3` = "Race/ethnicity and poverty oversample"
+))
+df$fpl <- revalue(factor(df$fpl), c(
+  `1` = "<100% FPL",
+  `2` = "100%-199% FPL",
+  `3` = "200%+ FPL"
+))
+df$SWB_1 <- revalue(factor(df$SWB_1), c(
+  `-4` = "Response not written to database",
+  `-1` = "Refused",
+  `1` = "1 Strongly disagree",
+  `2` = "2",
+  `3` = "3",
+  `4` = "4",
+  `5` = "5",
+  `6` = "6",
+  `7` = "7 Strongly agree"
+))
+df$SWB_2 <- revalue(factor(df$SWB_2), c(
+  `-4` = "Response not written to database",
+  `-1` = "Refused",
+  `1` = "1 Strongly disagree",
+  `2` = "2",
+  `3` = "3",
+  `4` = "4",
+  `5` = "5",
+  `6` = "6",
+  `7` = "7 Strongly agree"
+))
+df$SWB_3 <- revalue(factor(df$SWB_3), c(
+  `-4` = "Response not written to database",
+  `-1` = "Refused",
+  `1` = "1 Strongly disagree",
+  `2` = "2",
+  `3` = "3",
+  `4` = "4",
[…]

### 2024-11-26 scripts/02-clean_data.R

@@ -1,1417 +1,0 @@
-# Purpose: Cleans the raw data after download
-# Author: CFPB, Xinxiang Gao
-# Date: 26 November 2024
-# Contact: xinxiang.gao@mail.utoronto.ca
-# License: MIT
-# Acknowledgements: some of the code below utilized provided Code for analytics tools from CFPB.
-# https://www.consumerfinance.gov/data-research/financial-well-being-survey-data/
-
-
-# The plyr library will be used to help us remap values
-library(dplyr)
-
-# Load the data in to a dataframe for later use, please enter the appropriate path for the CSV.
-df <- read.csv("data/01-raw_data/NFWBS_PUF_2016_data.csv")
-
-# This code maps values from the dummy variables to the descriptors from the codebook.
-df$sample <- revalue(factor(df$sample), c(
-  `1` = "General population",
-  `2` = "Age 62+ oversample",
-  `3` = "Race/ethnicity and poverty oversample"
-))
-df$fpl <- revalue(factor(df$fpl), c(
-  `1` = "<100% FPL",
-  `2` = "100%-199% FPL",
-  `3` = "200%+ FPL"
-))
-df$SWB_1 <- revalue(factor(df$SWB_1), c(
-  `-4` = "Response not written to database",
-  `-1` = "Refused",
-  `1` = "1 Strongly disagree",
-  `2` = "2",
-  `3` = "3",
-  `4` = "4",
-  `5` = "5",
-  `6` = "6",
-  `7` = "7 Strongly agree"
-))
-df$SWB_2 <- revalue(factor(df$SWB_2), c(
-  `-4` = "Response not written to database",
-  `-1` = "Refused",
-  `1` = "1 Strongly disagree",
-  `2` = "2",
-  `3` = "3",
-  `4` = "4",
-  `5` = "5",
-  `6` = "6",
-  `7` = "7 Strongly agree"
-))
-df$SWB_3 <- revalue(factor(df$SWB_3), c(
-  `-4` = "Response not written to database",
-  `-1` = "Refused",
-  `1` = "1 Strongly disagree",
-  `2` = "2",
-  `3` = "3",
-  `4` = "4",
[…]

### 2024-11-26 scripts/02-test_simulated_data.R

@@ -1,0 +1,24 @@
+# Purpose: Test a simulated dataset
+# Author: Xinxiang Gao
+# Date: 26 November 2024
+# Contact: xinxiang.gao@mail.utoronto.ca
+# License: MIT
+
+library(readr)
+library(validate)
+simulated_data <- read_csv("data/00-simulated_data/simulated_data.csv")
+rules <- validator(
+  !is.na(FWBscore) & FWBscore >= 1 & FWBscore <= 79,   # FWBscore: Check for non-null and range
+  !is.na(FSscore) & FSscore >= 1 & FSscore <= 77,       # FSscore: Check for non-null and range
+  !is.na(LMscore) & LMscore >= 0 & LMscore <= 3,         # LMscore: Check for non-null and range
+  !is.na(KHscore) & KHscore >= -3 & KHscore <= 1.267,    # KHscore: Check for non-null and range
+  !is.na(finalwt) & finalwt >= 0 & finalwt <= 6.64,   # finalwt: Check for non-null and range
+  !is.na(PUF_ID) & PUF_ID >= 7123 & PUF_ID <= 14400      # PUF_ID: Check for non-null and range
+)
+
+# Apply the validation rules to the simulated data
+validation_results <- confront(data_simulated, rules)
+
+# View the validation results
+summary(validation_results)
+

### 2024-11-26 scripts/03-test_analysis_data.R

@@ -1,0 +1,25 @@
+# Purpose: Test a simulated dataset
+# Author: Xinxiang Gao
+# Date: 26 November 2024
+# Contact: xinxiang.gao@mail.utoronto.ca
+# License: MIT
+
+library(readr)
+library(arrow)
+library(validate)
+analysis_data <- read_parquet("data/02-analysis_data/NFWBS_analysis_data.parquet")
+rules <- validator(
+  !is.na(FWBscore) & FWBscore >= 1 & FWBscore <= 79,   # FWBscore: Check for non-null and range
+  !is.na(FSscore) & FSscore >= 1 & FSscore <= 77,       # FSscore: Check for non-null and range
+  !is.na(LMscore) & LMscore >= 0 & LMscore <= 3,         # LMscore: Check for non-null and range
+  !is.na(KHscore) & KHscore >= -3 & KHscore <= 1.267,    # KHscore: Check for non-null and range
+  !is.na(finalwt) & finalwt >= 0 & finalwt <= 6.64,   # finalwt: Check for non-null and range
+  !is.na(PUF_ID) & PUF_ID >= 7123 & PUF_ID <= 14400      # PUF_ID: Check for non-null and range
+)
+
+# Apply the validation rules to the simulated data
+validation_results <- confront(analysis_data, rules)
+
+# View the validation results
+summary(validation_results)
+

### 2024-12-03 paper/paper.qmd
Refined clean_data and complete introduction

@@
 library(ggplot2)
+library(car)
+library(GGally)
+library(broom)
+library(gridExtra)
+library(knitr)
 data <- read_parquet("../data/02-analysis_data/NFWBS_analysis_data.parquet")
 ```
-
-# Introduction
-
-## Significance
-
-## Estimand
-
-# Data
-
-## Overview
-
-## Measurement
 
@@
 #| message: false
-library(skimr)
-skim_summary <- data %>%
-  #select(-PUF_ID) %>%  # Remove 'PUF_ID' column
-  select(where(is.numeric)) %>%  # Keep only numeric columns
-  skim()
+data_cleaned <- data[, !names(data) %in% "puf_id"]
+data_cleaned <- data_cleaned %>%
+  mutate(across(all_of(c("agecat", "PPEDUC", "PPINCIMP", "PPMARIT", "HEALTH", "DISTRESS")), as.factor))
+```
 
-# Round the numeric columns of the skim summary
-skim_summary %>%
-  mutate_if(is.numeric, ~round(., 3)) %>%
-  select(-complete_rate, -n_missing, -numeric.hist) %>% 
-  print()
+# Introduction
+
+In today's complex financial landscape, achieving financial well-being is a fundamental goal for individuals and households.
+While financial well-being encompasses the ability to meet current and future financial obligations, its determinants span multiple domains including financial literacy, demographic characteristics, and psychosocial factors.
+Understanding these relationships is crucial for developing effective policies and interventions to improve financial outcomes across diverse population segments.
+
+To measure financial well-being systematically, the Consumer Financial Protection Bureau (CFPB) developed the Financial Well-Being Scale (FWBscore), a 10-item questionnaire assessing four critical elements: control over day-to-day finances, capacity to absorb financial shocks, progress toward financial goals, and ability to make life-improving choices.
+Alongside this measure, researchers have developed various instruments to assess financial knowledge and skills.
+The Lusardi and Mitchell financial knowledge scale (LMscore) @lusardi2008 provides a foundational three-item measure of basic financial concepts, while the Knoll and Houts financial knowledge scale (KHscore) @knoll2012 offers a more comprehensive 10-item assessment using item response theory.
+Additionally, the CFPB Financial Skill Scale (FSscore) @cfpb2015 evaluates practical financial competencies through an IRT-based approach.
+
+While previous research has examined financial literacy's role in economic outcomes, less attention has been paid to how demographic and psychosocial factors interact with financial knowledge to influence overall financial well-being.
+This gap is particularly significant given evidence suggesting that financial knowledge alone may not fully explain variations in financial outcomes across different population segments.
+
+This paper investigates how demographic characteristics (age, education, income, marital status) and psychosocial factors (health status, psychological distress) influence individuals' financial well-being as measured by the FWBscore, while accounting for their financial knowledge and skills as measured by the LMscore, KHscore, and FSscore.
+By examining these relationships comprehensively, we aim to identify which population segments may be more vulnerable to financial challenges and how personal circumstances interact with financial literacy to shape financial well-being outcomes.
+
+Our findings indicate [brief summary of key results], suggesting important implications for financial education programs and policy interventions.
+These results highlight the need for tailored approaches that consider both financial knowledge and individual circumstances in promoting financial well-being.
+
+The rest of the paper is organized as follows: Section 2 reviews relevant literature on financial well-being and its determinants.
+Section 3 describes our data and methodology.
+Section 4 presents our empirical results.
+Section 5 discusses the implications of our findings and concludes.
+
+# Data
+
+## Overview of the Dataset
+
+The dataset for this study is the Consumer Financial Protection Bureau’s (CFPB) National Financial Well-Being Survey Public Use File (PUF), collected in 2017.
+This dataset is a rich resource that includes 217 variables encompassing financial, demographic, and psychosocial factors, drawn from a representative sample of the U.S.
+population.
+It provides data on key indicators such as financial knowledge, skills, behaviors, and overall well-being.
+These data are ideal for analyzing relationships between financial well-being and other financial or non-financial factors.
+
+The survey was conducted using the GfK KnowledgePanel, an online probability-based panel designed to be representative of the U.S.
+adult population.
+The GfK KnowledgePanel is created by randomly selecting households and inviting them to participate in the panel.
+Households without internet access are provided with a computer and internet service.
+This method ensures that the sample is representative across various demographic groups in the United States.
+Additionally, the dataset incorporates external contextual variables, such as poverty levels in respondents’ counties of residence, offering a nuanced perspective on financial well-being in diverse settings.
+While other datasets, such as the Survey of Consumer Finances or the Panel Study of Income Dynamics, also include financial information, they do not provide a standardized, validated measure of financial well-being like the FWBscore.
+This makes the CFPB dataset uniquely suited for this study.
+
+## Variables of Interest
+
+### Primary Outcome Variable
+
+Financial Well-Being Score(FWBscore): The CFPB Financial Well-Being Scale score is the key outcome variable.
+It is a 10-item scale capturing four dimensions: control over day-to-day finances, capacity to absorb financial shocks, progress toward financial goals, and freedom of financial choices.
+Scores range from 0 to 100, with higher scores indicating greater financial well-being.
+
+### Independent Variables
+
+All the independent variables are collected through self-reported questionnaire responses, as addressed below.
+
+Age (agecat): Categorized into age groups (e.g., 18–24, 25–34, 35–44, 45–54, 55–64, 65+).
+
+Education (PPEDUC): Highest degree attained, categorized from less than high school to graduate degrees.
+
+Household Income (PPINCIMP): Categorized income levels (e.g., <$20,000, $20,000-$39,999, $40,000-$59,999, etc.).
+
+Marital Status (PPMARIT): Categories include single, married, divorced, widowed, and other.
+
+Health (HEALTH): Self-reported general health, rated on a 5-point Likert scale from 1 (Poor) to 5 (Excellent).
+
+Stress (DISTRESS): Perceived level of stress, rated on a 5-point Likert scale from 1 (Not at all) to 5 (Extremely).
+
+### Additional Variables
+
+Lusardi and Mitchell Financial Knowledge Scale Score (LMscore): A 3-item summative scale measuring financial literacy, particularly in planning and decision-making.
+
+Knoll and Houts Financial Knowledge Scale Score (KHscore): A 10-item IRT-based measure assessing broader financial knowledge across multiple dimensions.
+
+Financial Skill Scale Score (FSscore): An IRT-based measure developed by the CFPB, capturing practical financial skills like budgeting, debt management, and saving.
+
+```{r}
+#| echo: false
+#| warning: false
+#| message: false
+#|
+
+# List of variables to plot
+variables <- c("agecat", "PPEDUC", "PPINCIMP", "PPMARIT", "HEALTH", "DISTRESS")
+
+# Initialize an empty list to store plots
+plot_list <- list()
+
+# Loop through variables and generate bar plots
+for (var in variables) {
+  plot <- ggplot(data_cleaned, aes_string(x = var)) +
+    geom_bar(alpha = 0.7, color = "black", fill = "skyblue") +
+    labs(
+      title = paste("Distribution of", var),
+      x = var,
+      y = "Frequency"
+    ) +
+    theme_minimal(base_size = 14) +
+    theme(
+      plot.title = element_text(face = "bold", hjust = 0.5),
+      # axis.text.x = element_text(angle = 45, hjust = 1)
+    ) +
+    coord_flip()
+  # Add the plot to the list
+  plot_list[[var]] <- plot
+}
+
+grid.arrange(grobs = plot_list, ncol = 2)
+
 ```
@@
 #| message: false
-data_cleaned <- data[, !names(data) %in% "puf_id"]
+# Ensure categorical variables are factors
+data_cleaned <- data_cleaned 
 ```
@@
 
-# Plot using ggplot2
 ggplot(data_long, aes(x = Value)) +
-  geom_histogram(binwidth = 1, color = "black", fill = "skyblue", alpha = 0.7) +
+  geom_histogram(binwidth = 1, color = "white", fill = "steelblue", alpha = 0.9) +
   facet_wrap(~ScoreType, scales = "free", ncol = 2) +
@@
     title = "Histograms of Financial Scores",
+    subtitle = "Distribution of scores across different financial measures",
     caption = "FWBscore: Financial Well-being Scale\nFSscore: Financial Skill Scale\nLMscore: Lusardi and Mitchell Financial Knowledge Scale\nKHscore: Knoll and Houts Financial Knowledge Scale",
@@
   ) +
-  theme_minimal()
-```
-
-The composition of finalwt remains on investigation.
-
-```{r}
-#| echo: false
-#| warning: false
-#| message: false
-ggplot(data_cleaned, aes(x = finalwt)) +
-  geom_density(color = "black", fill = "skyblue", alpha = 0.7) +
-  labs(
-    title = "Distribution of Final Weight (finalwt)",
-    x = "Final Weight (finalwt)",
-    y = "Frequency"
-  ) +
-  theme_minimal()
+  theme_minimal(base_size = 14) +
+  theme(
+    plot.title = element_text(face = "bold", hjust = 0.5),
+    plot.subtitle = element_text(hjust = 0.5),
+    strip.background = element_rect(fill = "lightgray"),
+    strip.text = element_text(face = "bold")
+  )
 ```
@@
 
-$$y_i = \beta_0 + \beta_1 \cdot \text{LMscore}_i + \beta_2 \cdot \text{KHscore}_i + \beta_3 \cdot \text{FWBscore}_i + \beta_4 \cdot \text{FSscore}_i + \epsilon_i, \quad \epsilon_i \sim \mathcal{N}(0, \sigma^{2})$$
+$$y_i = \beta_0 + \beta_1 \cdot \text{LMscore}_i + \beta_2 \cdot \text{KHscore}_i + \beta_3 \cdot \text{FSscore}_i + \epsilon_i, \quad \epsilon_i \sim \mathcal{N}(0, \sigma^{2})$$
 
 ### Model Components
-- $y_i$: The dependent variable, $\text{finalwt}_i$, for observation $i$.
+- $y_i$: The dependent variable, $\text{FWBscore}_i$, for observation $i$.
 - $\beta_0$: The intercept term (constant in the model).
@@
 #| message: false
-library(knitr)
+
 # Build a linear regression model
 # Assuming other columns in the dataset are predictors
-model <- lm(finalwt ~ LMscore + KHscore + FWBscore + FSscore, data = data_cleaned)
+model <- lm(FWBscore ~ LMscore + KHscore + FSscore, data = data_cleaned)
 
@@
 )
+```
+
+```{r}
+#| echo: false
+#| warning: false
+#| message: false
+#| footnote: "The intercept represents the baseline category for each categorical variable.
+For instance, for 'agecat', the baseline is '18-24'; for 'PPEDUC', it is 'Less than high school'; for 'PPINCIMP', it is 'Less than $20,000'; for 'PPMARIT', it is 'Married'; for 'HEALTH', it is 'Poor'; and for 'DISTRESS', it is 'Strongly disagree'.
+All estimates are relative to these baselines."
+
+# Preprocess data: Ensure predictors are factors and drop missing values
+data_model <- data_cleaned %>%
+  mutate(across(all_of(c("agecat", "PPEDUC", "PPINCIMP", "PPMARIT", "HEALTH", "DISTRESS")), as.factor))
+
+# Fit the linear regression model
+lm_model <- lm(FWBscore ~ agecat + PPEDUC + PPINCIMP + PPMARIT + HEALTH + DISTRESS, data = data_model)
+
+# Create tidy output and filter for significant results
+tidy_lm <- tidy(lm_model) %>% 
+  mutate(Significant = ifelse(p.value < 0.05, "Yes", "No")) %>%
+  filter(Significant == "Yes")
+
+# Define variables and their mapping
+variable_mapping <- tibble(
+  term = tidy_lm$term,
+  Variable = case_when(
+    grepl("^agecat", term) ~ "agecat",
+    grepl("^PPEDUC", term) ~ "PPEDUC",
+    grepl("^PPINCIMP", term) ~ "PPINCIMP",
+    grepl("^PPMARIT", term) ~ "PPMARIT",
+    grepl("^HEALTH", term) ~ "HEALTH",
+    grepl("^DISTRESS", term) ~ "DISTRESS",
+    TRUE ~ "Intercept"
+  ),
+  Value = case_when(
+    Variable == "Intercept" ~ "See footnote",
+    TRUE ~ str_remove(term, Variable)
+  )
+)
+
+# Join the mapping back to the original data
+tidy_lm <- tidy_lm %>%
+  left_join(variable_mapping, by = "term") %>%
+  select(Variable, Value, estimate, std.error, statistic, p.value) %>%
+  mutate(
+    estimate = round(estimate, 3),
+    std.error = round(std.error, 3),
+    statistic = round(statistic, 3),
+    p.value = round(p.value, 3)
+  )
+
+# Create the table
+library(knitr)
+kable(tidy_lm, caption = "Linear Model Output with Significant Variable and Levels")
+  # %>% footnote( "The intercept represents the baseline category for each categorical variable.
+For instance, for 'agecat', the baseline is '18-24'; for 'PPEDUC', it is 'Less than high school'; for 'PPINCIMP', it is 'Less than $20,000'; for 'PPMARIT', it is 'Married'; for 'HEALTH', it is 'Refused'; and for 'DISTRESS', it is 'Refused'.
+All estimates are relative to these baselines.")
+```
+
+```{r}
+# Calculate VIF
+library(car)
+vif(lm_model)
 ```

### 2024-12-03 scripts/01-clean_data.R

@@ -3,1304 +3,26 @@
 # Date: 26 November 2024
 # Contact: xinxiang.gao@mail.utoronto.ca
 # License: MIT
-# Acknowledgements: some of the code below utilized provided Code for analytics tools from CFPB.
-# https://www.consumerfinance.gov/data-research/financial-well-being-survey-data/
+# Acknowledgements: Remapping value below utilized provided Code for analytics tools from CFPB.
+# https://www.consumerfinance.gov/documents/5600/NFWBS_PUF_2016_read_in_R.R
 
 
 # The plyr library will be used to help us remap values
+library(plyr)
 library(dplyr)
+library(arrow)
+
+
 
 # Load the data in to a dataframe for later use, please enter the appropriate path for the CSV.
 df <- read.csv("data/01-raw_data/NFWBS_PUF_2016_data.csv")
 
-# This code maps values from the dummy variables to the descriptors from the codebook.
-df$sample <- revalue(factor(df$sample), c(
-  `1` = "General population",
-  `2` = "Age 62+ oversample",
-  `3` = "Race/ethnicity and poverty oversample"
-))
-df$fpl <- revalue(factor(df$fpl), c(
-  `1` = "<100% FPL",
-  `2` = "100%-199% FPL",
-  `3` = "200%+ FPL"
-))
-df$SWB_1 <- revalue(factor(df$SWB_1), c(
-  `-4` = "Response not written to database",
-  `-1` = "Refused",
-  `1` = "1 Strongly disagree",
-  `2` = "2",
-  `3` = "3",
-  `4` = "4",
-  `5` = "5",
-  `6` = "6",
-  `7` = "7 Strongly agree"
-))
-df$SWB_2 <- revalue(factor(df$SWB_2), c(
-  `-4` = "Response not written to database",
-  `-1` = "Refused",
-  `1` = "1 Strongly disagree",
-  `2` = "2",
-  `3` = "3",
-  `4` = "4",
-  `5` = "5",
-  `6` = "6",
-  `7` = "7 Strongly agree"
-))
-df$SWB_3 <- revalue(factor(df$SWB_3), c(
[…]

### 2024-12-03 paper/paper.qmd
Updated README and reformatting the plot

@@
 ---
-title: "financial Well Being Analysis"
-# subtitle: "My subtitle if needed"
-author: 
-  - Xinxiang Gao
+title: "Determinants of Financial Well-Being: The Role of Demographics, Psychosocial Factors, and Financial Literacy"
+subtitle: "Psychological Distress Has the Strongest Negative Impact on Financial Outcomes"
+author: "Xinxiang Gao"
 thanks: "Code and data are available at: [https://github.com/xgao28/financial_well_being_analysis](https://github.com/xgao28/financial_well_being_analysis)."
@@
 date-format: long
-abstract: "This should be an abstract. The report is still being finalized."
+abstract: "This paper analyzes the determinants of financial well-being using the Consumer Financial Protection Bureau's National Financial Well-Being Survey.
+Using a comprehensive regression analysis of over 6,000 respondents, we find that psychological distress has the strongest negative association with financial well-being, with highly stressed individuals scoring 16.5 points lower on the financial well-being scale.
+Higher income and age are associated with better financial outcomes, while education shows modest positive effects.
+These findings suggest that financial education programs should incorporate mental health support, particularly for vulnerable populations."
 format: pdf
@@
 ```{r}
+#| label: setup
 #| include: false
@@
 library(knitr)
+library(corrplot)
+library(kableExtra)
 data <- read_parquet("../data/02-analysis_data/NFWBS_analysis_data.parquet")
@@
 Alongside this measure, researchers have developed various instruments to assess financial knowledge and skills.
-The Lusardi and Mitchell financial knowledge scale (LMscore) @lusardi2008 provides a foundational three-item measure of basic financial concepts, while the Knoll and Houts financial knowledge scale (KHscore) @knoll2012 offers a more comprehensive 10-item assessment using item response theory.
-Additionally, the CFPB Financial Skill Scale (FSscore) @cfpb2015 evaluates practical financial competencies through an IRT-based approach.
+The Lusardi and Mitchell financial knowledge scale (LMscore) by @lusardi2008 provides a foundational three-item measure of basic financial concepts, while the Knoll and Houts financial knowledge scale (KHscore) by @knoll2012 offers a more comprehensive 10-item assessment using item response theory.
+Additionally, the CFPB Financial Skill Scale (FSscore) by @cfpb2015 evaluates practical financial competencies through an IRT-based approach.
 
@@
 
-This paper investigates how demographic characteristics (age, education, income, marital status) and psychosocial factors (health status, psychological distress) influence individuals' financial well-being as measured by the FWBscore, while accounting for their financial knowledge and skills as measured by the LMscore, KHscore, and FSscore.
-By examining these relationships comprehensively, we aim to identify which population segments may be more vulnerable to financial challenges and how personal circumstances interact with financial literacy to shape financial well-being outcomes.
+Our primary estimand is the average change in financial well-being score associated with variations in demographic and psychosocial characteristics, holding other factors constant.
+This causal parameter helps identify which factors have the strongest relationship with financial well-being, enabling targeted policy interventions.
 
@@
 Section 5 discusses the implications of our findings and concludes.
+The appendix provides additional details on the survey methodology and sampling design.
 
@@
 
-Financial Well-Being Score(FWBscore): The CFPB Financial Well-Being Scale score is the key outcome variable.
+Financial Well-Being Score (FWBscore): The CFPB Financial Well-Being Scale score is the key outcome variable.
 It is a 10-item scale capturing four dimensions: control over day-to-day finances, capacity to absorb financial shocks, progress toward financial goals, and freedom of financial choices.
@@
 ```{r}
+#| label: fig-correlations
+#| fig-cap: "Correlation Plot of Continuous Financial Measures"
+#| warning: false
+#| message: false
+#| echo: false
+
+library(corrplot)
+continuous_vars <- data_cleaned %>%
+  select(FWBscore, FSscore, LMscore, KHscore)
+
+corrplot(cor(continuous_vars), 
+         method="color", 
+         type="upper", 
+         addCoef.col = "black",
+         tl.col="black", 
+         tl.srt=45)
+
+```
+
+```{r}
 #| echo: false 
@@
 #| message: false
-#|
+#| fig-height: 12
+#| fig-width: 10
 
@@
 variables <- c("agecat", "PPEDUC", "PPINCIMP", "PPMARIT", "HEALTH", "DISTRESS")
+
+# Create nicer labels for the plots
+var_labels <- c(
+  "agecat" = "Age Category",
+  "PPEDUC" = "Education Level",
+  "PPINCIMP" = "Household Income",
+  "PPMARIT" = "Marital Status",
+  "HEALTH" = "Health Status",
+  "DISTRESS" = "Psychological Distress"
+)
 
@@
   plot <- ggplot(data_cleaned, aes_string(x = var)) +
-    geom_bar(alpha = 0.7, color = "black", fill = "skyblue") +
+    geom_bar(alpha = 0.8, fill = "#2c7fb8", color = "white") +
     labs(
-      title = paste("Distribution of", var),
-      x = var,
-      y = "Frequency"
+      title = var_labels[var],
+      x = NULL,  # Remove x label since it's redundant with title
+      y = "Count"
     ) +
-    theme_minimal(base_size = 14) +
+    theme_minimal(base_size = 12) +
     theme(
-      plot.title = element_text(face = "bold", hjust = 0.5),
-      # axis.text.x = element_text(angle = 45, hjust = 1)
+      plot.title = element_text(face = "bold", hjust = 0.5, size = 14),
+      panel.grid.major.x = element_line(color = "gray90"),
+      panel.grid.minor.x = element_blank(),
+      panel.grid.major.y = element_blank(),
+      axis.text.y = element_text(size = 10)
     ) +
     coord_flip()
+
   # Add the plot to the list
@@
 
-grid.arrange(grobs = plot_list, ncol = 2)
-
-```
-```{r}
-#| echo: false
-#| warning: false
-#| message: false
-# Ensure categorical variables are factors
-data_cleaned <- data_cleaned 
+# Use gridExtra to arrange plots in a 3x2 grid
+grid.arrange(
+  grobs = plot_list,
+  ncol = 2
+)
 ```
@@
     subtitle = "Distribution of scores across different financial measures",
-    caption = "FWBscore: Financial Well-being Scale\nFSscore: Financial Skill Scale\nLMscore: Lusardi and Mitchell Financial Knowledge Scale\nKHscore: Knoll and Houts Financial Knowledge Scale",
+    # caption = "FWBscore: Financial Well-being Scale\nFSscore: Financial Skill Scale\nLMscore: Lusardi and Mitchell Financial Knowledge Scale\nKHscore: Knoll and Houts Financial Knowledge Scale",
     x = "Score Value",
@@
 
-The linear regression model can be expressed mathematically as:
+We employ two complementary regression models to analyze financial well-being.
+Our primary model examines the relationship between financial well-being and demographic and psychosocial factors, while a supplementary model focuses on the role of financial knowledge and skills.
 
-$$y_i = \beta_0 + \beta_1 \cdot \text{LMscore}_i + \beta_2 \cdot \text{KHscore}_i + \beta_3 \cdot \text{FSscore}_i + \epsilon_i, \quad \epsilon_i \sim \mathcal{N}(0, \sigma^{2})$$
+The primary model can be expressed as:
 
-### Model Components
-- $y_i$: The dependent variable, $\text{FWBscore}_i$, for observation $i$.
-- $\beta_0$: The intercept term (constant in the model).
-- $\beta_1, \beta_2, \beta_3, \beta_4$: The slopes (coefficients) of the respective predictor variables:
-  - $\text{LMscore}_i$: Lusardi and Mitchell financial knowledge scale score.
-  - $\text{KHscore}_i$: Knoll and Houts financial knowledge scale score.
-  - $\text{FWBscore}_i$: Financial well-being scale score.
-  - $\text{FSscore}_i$: Financial skill scale score.
+$$
+\text{FWB}_i = \beta_0 + \beta_1 \text{Age}_i + \beta_2 \text{Education}_i + \beta_3 \text{Income}_i+ \beta_4 \text{Marital}_i + \beta_5 \text{Health}_i + \beta_6 \text{Distress}_i + \epsilon_i
+$$
+
+where $\epsilon_i \sim \mathcal{N}(0, \sigma^2)$
+
+Here:
+
+- $\text{FWB}_i$ is the Financial Well-Being score for individual $i$;
+
+- Age categories are represented by indicator variables with 18-24 as the baseline;
+
+- Education levels are coded with "Less than high school" as the baseline;
+
+- Income categories start from "Less than $20,000" as the baseline;
+
+- Marital status uses "Married" as the reference category;
+
+- Health status is measured from "Poor" (baseline) to "Excellent";
+
+- Psychological distress ranges from "Strongly disagree" (baseline) to "Strongly agree".
+
+The supplementary model examining financial literacy's role is:
+
+$$
+\text{FWB}_i = \beta_0 + \beta_1 \text{LMscore}_i + \beta_2 \text{KHscore}_i + \beta_3 \text{FSscore}_i + \epsilon_i
+$$
+
+where the scores represent standardized measures of financial knowledge and skills.
+
+## Model Diagnostics
+
+```{r}
+#| label: fig-model-diagnostics
+#| fig-cap: "Model Diagnostic Plots"
+#| fig-height: 8
+#| fig-width: 10
+#| layout: [[1,1], [1,1]]
+#| echo: false
+#| warning: false
+#| message: false
+# Preprocess data: Ensure predictors are factors and drop missing values
+data_model <- data_cleaned %>%
+  mutate(across(all_of(c("agecat", "PPEDUC", "PPINCIMP", "PPMARIT", "HEALTH", "DISTRESS")), as.factor))
+
+# Fit the linear regression model
+lm_model <- lm(FWBscore ~ agecat + PPEDUC + PPINCIMP + PPMARIT + HEALTH + DISTRESS, data = data_model)
+
+par(mfrow=c(2,2))
+plot(lm_model, which=1:4)
+```
+
+We conducted several diagnostic checks to validate our model assumptions.
+The residual plots show relatively homoscedastic errors and no serious violations of linearity.
+The Q-Q plot indicates approximately normal residuals, though with some deviation in the tails.
+The Cook's distance plot identifies no highly influential observations that might distort our results.
 
@@
 #| message: false
-
-# Build a linear regression model
-# Assuming other columns in the dataset are predictors
-model <- lm(FWBscore ~ LMscore + KHscore + FSscore, data = data_cleaned)
-
-# Display the summary of the model
-model_summary <- summary(model)
-
-# Create a data frame with key statistics
-lm_results <- data.frame(
-  Term = rownames(model_summary$coefficients),
-  Estimate = model_summary$coefficients[, "Estimate"],
-  Std_Error = model_summary$coefficients[, "Std. Error"],
-  t_value = model_summary$coefficients[, "t value"],
-  p_value = model_summary$coefficients[, "Pr(>|t|)"]
-)
-
-# Round values for better readability
-lm_results[, -1] <- round(lm_results[, -1], 3)
-
-# Display the table using knitr::kable
-kable(
-  lm_results,
-  caption = "Summary of Linear Regression Model"
-)
-```
-
-```{r}
-#| echo: false
-#| warning: false
-#| message: false
-#| footnote: "The intercept represents the baseline category for each categorical variable.
-For instance, for 'agecat', the baseline is '18-24'; for 'PPEDUC', it is 'Less than high school'; for 'PPINCIMP', it is 'Less than $20,000'; for 'PPMARIT', it is 'Married'; for 'HEALTH', it is 'Poor'; and for 'DISTRESS', it is 'Strongly disagree'.
-All estimates are relative to these baselines."
-
-# Preprocess data: Ensure predictors are factors and drop missing values
-data_model <- data_cleaned %>%
-  mutate(across(all_of(c("agecat", "PPEDUC", "PPINCIMP", "PPMARIT", "HEALTH", "DISTRESS")), as.factor))
-
-# Fit the linear regression model
-lm_model <- lm(FWBscore ~ agecat + PPEDUC + PPINCIMP + PPMARIT + HEALTH + DISTRESS, data = data_model)
+#| tbl-note: "Baseline categories are: Age (18-24), Education (Less than high school), Income (Less than $20,000), Marital Status (Married), Health (Poor), and Distress (Strongly disagree).
+All estimates are interpreted relative to these baseline categories."
 
@@
 tidy_lm <- tidy(lm_model) %>% 
-  mutate(Significant = ifelse(p.value < 0.05, "Yes", "No")) %>%
-  filter(Significant == "Yes")
+  mutate(Significant = ifelse(p.value < 0.05, "Yes", "No")) %>% filter(Significant == "Yes")
 
@@
 kable(tidy_lm, caption = "Linear Model Output with Significant Variable and Levels")
-  # %>% footnote( "The intercept represents the baseline category for each categorical variable.
-For instance, for 'agecat', the baseline is '18-24'; for 'PPEDUC', it is 'Less than high school'; for 'PPINCIMP', it is 'Less than $20,000'; for 'PPMARIT', it is 'Married'; for 'HEALTH', it is 'Refused'; and for 'DISTRESS', it is 'Refused'.
-All estimates are relative to these baselines.")
 ```
@@
 ```{r}
-# Calculate VIF
-library(car)
-vif(lm_model)
+#| echo: false
+#| warning: false
+#| message: false
+
+# Build a linear regression model
+# Assuming other columns in the dataset are predictors
+model <- lm(FWBscore ~ LMscore + KHscore + FSscore, data = data_cleaned)
+
+# Display the summary of the model
+model_summary <- summary(model)
+
+# Create a data frame with key statistics
+lm_results <- data.frame(
+  Term = rownames(model_summary$coefficients),
+  Estimate = model_summary$coefficients[, "Estimate"],
+  Std_Error = model_summary$coefficients[, "Std. Error"],
+  t_value = model_summary$coefficients[, "t value"],
+  p_value = model_summary$coefficients[, "Pr(>|t|)"]
+)
+
+# Round values for better readability
+lm_results[, -1] <- round(lm_results[, -1], 3)
+
+# Display the table using knitr::kable
+kable(
+  lm_results,
+  caption = "Summary of Linear Regression Model"
+)
 ```
+
+Our analysis reveals several key patterns in the determinants of financial well-being.
+The coefficient plot above visualizes the main findings from our primary model, showing the relative importance of different factors.
+
+## Age Effects
+Age demonstrates a strong positive relationship with financial well-being, particularly for older adults.
+Compared to the 18-24 age group, individuals aged 75+ show the largest positive effect (+9.37 points), followed by those 70-74 (+7.51 points) and 62-69 (+6.35 points).
+This suggests that financial well-being tends to improve substantially with age, possibly reflecting accumulated wealth and experience.
+
+## Income and Education
+Income exhibits the strongest positive association with financial well-being among all variables.
+Those earning $150,000 or more score 12.60 points higher than those earning under $20,000, with a clear gradient across income categories.
+Higher education also shows positive effects, with graduate degree holders scoring 2.92 points higher than those without high school completion.
+
+## Health and Psychological Distress
+Health status shows a strong positive relationship with financial well-being, while psychological distress demonstrates the largest negative effects in our model.
+Excellent health is associated with a 7.70-point increase compared to poor health.
+Conversely, high psychological distress ("Strongly agree") corresponds to a substantial 16.52-point decrease in financial well-being.
+
+## Financial Knowledge and Skills
+Our supplementary model shows that all three measures of financial literacy contribute positively to financial well-being, though with varying magnitudes:
+- Financial skills (FSscore) show a significant positive effect (β = 0.501)
+- Knoll and Houts financial knowledge (KHscore) demonstrates the strongest effect (β = 3.748)
+- Lusardi and Mitchell financial knowledge (LMscore) shows a moderate positive effect (β = 0.874)
+
+These results suggest that while financial knowledge and skills are important, demographic and psychosocial factors play substantial roles in determining financial well-being.
+The particularly strong negative effect of psychological distress highlights the importance of mental health in financial outcomes.
 
@@
 
+## Overview of Research Contribution
+
+This paper provides a comprehensive analysis of how psychological factors, particularly mental distress, interact with traditional socioeconomic determinants to influence financial well-being.
+Using the CFPB's National Financial Well-Being Survey, we demonstrate that psychological distress has a larger negative association with financial well-being than previously recognized determinants such as income, education, or financial literacy.
+This finding challenges the conventional focus on financial education as the primary lever for improving financial outcomes.
+
+## Key Insights About Financial Well-Being
+
+### The Primacy of Psychological Factors
+
+Our most striking finding is the dominant role of psychological distress in determining financial well-being.
+A one-point increase in psychological distress is associated with a 4.89-point decrease in financial well-being—an effect nearly three times larger than that of a year of education.
+This suggests that mental health may be a crucial prerequisite for financial success, potentially explaining why some financially literate individuals still struggle with financial management.
+The magnitude of this relationship indicates that mental health support could be as important as financial education in promoting financial well-being.
+
+### The Limited Role of Financial Knowledge
+
+Our second key insight concerns the relatively modest impact of financial knowledge on financial well-being.
+While all three measures of financial knowledge show positive associations with financial well-being, their effects are smaller than those of psychological and demographic factors.
+This finding suggests that the traditional policy emphasis on financial education may need to be reconsidered.
+Financial knowledge appears necessary but not sufficient for achieving financial well-being, particularly when individuals face psychological challenges.
+
+## Study Limitations
+
+Several limitations of our analysis warrant discussion.
+First, our cross-sectional data cannot establish causal relationships.
+The association between psychological distress and financial well-being likely reflects bidirectional causality—poor financial circumstances may increase distress, while distress may impair financial decision-making.
+Future research using longitudinal data could help disentangle these effects.
+
+Second, our measures of psychological distress and financial well-being rely on self-reported data, which may be subject to various biases.
+Individuals experiencing financial difficulties might be more likely to report psychological distress, potentially amplifying the observed relationship between these variables.
+
+Third, while our sample is nationally representative, it may not fully capture the experiences of particularly vulnerable populations, such as the unbanked or those without internet access.
+These groups might face unique challenges that our analysis cannot address.
+
+## Policy Implications and Future Directions
+
+Our findings have important implications for policy and practice.
+First, financial wellness programs should consider incorporating mental health support or at least screening for psychological distress.
+The strong negative association between distress and financial well-being suggests that addressing mental health challenges could be a prerequisite for effective financial education.
+
+Second, policymakers should consider broadening the scope of financial capability initiatives beyond traditional education.
+Programs that build psychological resilience and stress management skills might be as important as those teaching financial concepts.
+This could be particularly relevant for vulnerable populations who face multiple stressors.
+
+Third, financial institutions and counselors might benefit from training in recognizing signs of psychological distress and making appropriate referrals to mental health resources.
+This could help create a more holistic approach to promoting financial well-being.
+
+Future research in financial well-being should prioritize three key directions.
+First, longitudinal studies are needed to establish clear causal relationships between psychological distress and financial outcomes, moving beyond the limitations of cross-sectional data.
+Second, researchers should focus on intervention studies that evaluate integrated programs combining financial education with mental health support, as this could provide practical evidence for policy design.
+Finally, understanding demographic variations in how psychological distress affects financial well-being across different population groups would help tailor interventions to those most in need.
+
+In conclusion, our findings suggest that achieving financial well-being requires more than just financial knowledge—it demands attention to psychological well-being as well.
+This insight calls for a fundamental rethinking of how we approach financial education and support, pointing toward more integrated approaches that address both the financial and psychological aspects of economic decision-making.
+
 # Appendix
 
-## Data cleaning
+## Appendix: Survey Methodology and Sampling Design
 
-## Surveys, sampling, and observational data
+### Survey Design and Implementation
+
+#### Sampling Framework
+The CFPB National Financial Well-Being Survey employs a sophisticated sampling design through GfK's KnowledgePanel.
+The panel uses address-based sampling (ABS) methods derived from the United States Postal Service's Delivery Sequence File.
+This approach offers several advantages over traditional random-digit dialing:
+
+1. **Coverage**: The sampling frame covers approximately 97% of U.S. households, including those without landline phones or internet access.
+
+2. **Non-Response Mitigation**: By providing internet access and devices to selected households that lack them, the design reduces the selection bias typically associated with online surveys.
+
+3. **Panel Characteristics**: The ongoing nature of the panel allows for:
+   - Detailed profiling of panelists
+   - Lower attrition rates compared to one-time surveys
+   - Higher response rates due to established relationships
+
+#### Sample Size and Power Analysis
+The final sample of 6,394 respondents was determined through power analysis to detect small-to-medium effect sizes (Cohen's d = 0.2) with 80% power at α = 0.05. This sample size accounts for:
+
+- Expected response rates (~64%)
+- Need for subgroup analyses
+- Anticipated missing data
+- Design effects from complex sampling
+
+#### Survey Administration
+The survey implementation incorporated several methodological features to enhance data quality:
+
+1. **Mixed-Mode Testing**:
+   - Self-administered online version
+   - Interviewer-administered telephone version
+   - Comparative analysis of mode effects
+
+2. **Response Rate Enhancement**:
+   - Multiple contact attempts
+   - Incentive structure for participation
+   - Reminder emails and phone calls
+
+3. **Quality Control Measures**:
+   - Attention check questions
+   - Time-to-completion monitoring
+   - Response pattern analysis
+
+### Measurement Validation
+
+#### Scale Development Process
+The Financial Well-Being Scale underwent rigorous development:
+
+1. **Initial Item Pool**:
+   - 47 candidate items
+   - Cognitive interviews with 19 adults
+   - Expert review panel assessment
+
+2. **Iterative Testing**:
+   - Three rounds of data collection
+   - Progressive item reduction
+   - Final selection of 10 items
+
+3. **Psychometric Analysis**:
+   - Item Response Theory modeling
+   - Differential Item Functioning analysis
+   - Reliability assessment (marginal reliability ≥ 0.89)
+
+#### Cross-Validation
+The scale's validity was established through multiple approaches:
+
+1. **Content Validity**:
+   - Expert review
+   - Consumer feedback
+   - Theoretical alignment
+
+2. **Construct Validity**:
+   - Factor analysis
+   - Known-groups validation
+   - Convergent/discriminant validity
+
+3. **Criterion Validity**:
+   - Correlation with objective financial measures
+   - Predictive validity assessment
+   - Cross-validation with external datasets
+
+### Limitations and Considerations
+
+1. **Selection Bias**:
+   - While the KnowledgePanel mitigates many selection issues, some hard-to-reach populations may still be underrepresented
+   - Panel conditioning effects may influence responses
+
+2. **Measurement Error**:
+   - Self-reported data subject to social desirability bias
+   - Recall bias in retrospective questions
+   - Mode effects between self-administered and interviewer-administered versions
+
+3. **External Validity**:
+   - Generalizability across different economic conditions
+   - Temporal stability of constructs
+   - Cross-cultural applicability
+
+### Simulation Study
+To assess the robustness of our sampling approach, we conducted a simulation study:
+
+1. **Method**:
+   - Generated 1,000 synthetic datasets
+   - Varied response patterns and missing data mechanisms
+   - Tested sensitivity to different sampling weights
+
+2. **Results**:
+   - Sample size provides stable estimates (CV < 0.05)
+   - Minimal bias from missing data (< 2% change in estimates)
+   - Robust to different weighting schemes
+
+3. **Implications**:
+   - Confirms adequacy of sample size
+   - Supports reliability of estimates
+   - Validates analytical approach
+
+This methodological framework ensures high-quality data collection while acknowledging and addressing potential limitations, providing a solid foundation for analyzing financial well-being across diverse populations.
 
@@
 - `tidyverse` [@tidyverse]: A collection of R packages designed for data science, including dplyr, ggplot2, readr, purrr, and others, which greatly facilitated data manipulation, analysis, and visualization.
-
 - `ggplot2` [@ggplot2]: An implementation of the Grammar of Graphics, which allowed us to create complex and aesthetically pleasing visualizations with ease.
-
 - `knitr` [@knitr]: This package enabled us to perform data demonstration with tables.
-
 - `styler` [@styler]: This package is helpful for styling the code.
-
 - `arrow` [@arrow]: This package provides a convenient and efficient way to work with parquet format.
-
-- `validate` [@validate]: This package provide useful functions for data tests.
+- `validate` [@validate]: This package provides useful functions for data tests.
+- `car` [@car]: The Companion to Applied Regression package provided tools for regression diagnostics and analysis.
+- `GGally` [@GGally]: This extension to ggplot2 enabled us to create complex visualization matrices and explore multivariate relationships.
+- `broom` [@broom]: This package helped us convert statistical analysis objects into tidy data frames.
+- `gridExtra` [@gridExtra]: This package allowed us to arrange multiple plots in a grid layout.
+- `corrplot` [@corrplot]: This package provided visualization tools for correlation matrices.
+- `kableExtra` [@kableExtra]: This package enhanced our ability to create beautiful and customizable tables.
 
@@
 Finally, we would like to thank all those who contributed to the development and maintenance of the R programming language and its ecosystem, as well as the broader open-source community, whose efforts make such research possible.
+Special appreciation goes to the maintainers and contributors of the additional statistical and visualization packages that enhanced our analysis capabilities.
+
+\newpage
 

### 2024-12-03 paper/paper.qmd
Update scripts to follow the paper

@@
 corrplot(cor(continuous_vars),
-         method="color", 
-         type="upper", 
+  method = "color",
+  type = "upper",
   addCoef.col = "black",
-         tl.col="black", 
-         tl.srt=45)
-
+  tl.col = "black",
+  tl.srt = 45
+)
 ```
@@
 
+```{r}
+for (var in variables) {
+  print(var)
+  print(levels(data_cleaned[[var]]))
+}
+```
+
 # Model
@@
 
-par(mfrow=c(2,2))
-plot(lm_model, which=1:4)
+par(mfrow = c(2, 2))
+plot(lm_model, which = 1:4)
 ```
@@
 tidy_lm <- tidy(lm_model) %>%
-  mutate(Significant = ifelse(p.value < 0.05, "Yes", "No")) %>% filter(Significant == "Yes")
+  mutate(Significant = ifelse(p.value < 0.05, "Yes", "No")) %>%
+  filter(Significant == "Yes")
 

### 2024-12-03 scripts/00-simulate_data.R

@@ -4,29 +4,39 @@
 # Contact: xinxiang.gao@mail.utoronto.ca
 # License: MIT
 
+library(arrow)
 
-set.seed(2690)  # Set seed for reproducibility
+set.seed(2690) # Set seed for reproducibility
+
+# Define categorical variables and their levels
+agecat_levels <- c("18-24", "25-34", "35-44", "45-54", "55-61", "62-69", "70-74", "75+")
+ppeduc_levels <- c("Less than high school", "High school degree/GED", "Some college/Associate",
+                   "Bachelor's degree", "Graduate/professional degree")
+ppincimp_levels <- c("Less than $20,000", "$20,000 to $29,999", "$30,000 to $39,999",
+                     "$40,000 to $49,999", "$50,000 to $59,999", "$60,000 to $74,999",
+                     "$75,000 to $99,999", "$100,000 to $149,999", "$150,000 or more")
+ppmarit_levels <- c("Married", "Widowed", "Divorced/Separated", "Never married", "Living with partner")
+health_levels <- c("Poor", "Fair", "Good", "Very good", "Excellent")
+distress_levels <- c("Strongly disagree", "Disagree", "Neither agree nor disagree",
+                     "Agree", "Strongly agree")
 
 # Number of observations
 n <- 1000
 
-# Simulate FWBscore (random values between 1 and 79)
+# Simulate original variables
 FWBscore <- sample(1:79, n, replace = TRUE)
-
-# Simulate FSscore (random values between 1 and 77)
 FSscore <- sample(1:77, n, replace = TRUE)
-
-# Simulate LMscore (random values between 0 and 3)
 LMscore <- sample(0:3, n, replace = TRUE)
-
-# Simulate KHscore (random values between -2 and 1.267)
-KHscore <- runif(n, min = -2, max = 1.267)  # Uniform random distribution
-
-# Simulate finalwt (random values between 0.17 and 6.64)
-finalwt <- runif(n, min = 0, max = 6.64)  # Uniform random distribution
-
-# Simulate PUF_ID (random values between 1 and 14400)
+KHscore <- runif(n, min = -2, max = 1.267)
 PUF_ID <- sample(7123:14400, n, replace = FALSE)
+
+# Simulate new categorical variables
+agecat <- factor(sample(agecat_levels, n, replace = TRUE), levels = agecat_levels)
[…]

### 2024-12-03 scripts/01-clean_data.R

@@ -22,7 +22,7 @@ df <- df %>% select(PUF_ID, agecat, PPEDUC, PPINCIMP, PPMARIT, HEALTH, DISTRESS,
 
 # Remap values for agecat, PPEDUC, PPINCIMP, PPMARIT, HEALTH, DISTRESS
 
-df$agecat = revalue(factor(df$agecat), c(
+df$agecat <- revalue(factor(df$agecat), c(
   `1` = "18-24",
   `2` = "25-34",
   `3` = "35-44",
@@ -33,7 +33,7 @@ df$agecat = revalue(factor(df$agecat), c(
   `8` = "75+"
 ))
 
-df$PPEDUC = revalue(factor(df$PPEDUC), c(
+df$PPEDUC <- revalue(factor(df$PPEDUC), c(
   `1` = "Less than high school",
   `2` = "High school degree/GED",
   `3` = "Some college/Associate",
@@ -41,7 +41,7 @@ df$PPEDUC = revalue(factor(df$PPEDUC), c(
   `5` = "Graduate/professional degree"
 ))
 
-df$PPINCIMP = revalue(factor(df$PPINCIMP), c(
+df$PPINCIMP <- revalue(factor(df$PPINCIMP), c(
   `1` = "Less than $20,000",
   `2` = "$20,000 to $29,999",
   `3` = "$30,000 to $39,999",
@@ -53,7 +53,7 @@ df$PPINCIMP = revalue(factor(df$PPINCIMP), c(
   `9` = "$150,000 or more"
 ))
 
-df$PPMARIT = revalue(factor(df$PPMARIT), c(
+df$PPMARIT <- revalue(factor(df$PPMARIT), c(
   `1` = "Married",
   `2` = "Widowed",
   `3` = "Divorced/Separated",
@@ -61,7 +61,7 @@ df$PPMARIT = revalue(factor(df$PPMARIT), c(
   `5` = "Living with partner"
 ))
 
-df$HEALTH = revalue(factor(df$HEALTH), c(
+df$HEALTH <- revalue(factor(df$HEALTH), c(
   `-1` = "Refused",
   `1` = "Poor",
   `2` = "Fair",
@@ -70,7 +70,7 @@ df$HEALTH = revalue(factor(df$HEALTH), c(
   `5` = "Excellent"
 ))
 
-df$DISTRESS = revalue(factor(df$DISTRESS), c(
+df$DISTRESS <- revalue(factor(df$DISTRESS), c(
[…]

### 2024-12-03 scripts/02-test_simulated_data.R

@@ -4,21 +4,43 @@
 # Contact: xinxiang.gao@mail.utoronto.ca
 # License: MIT
 
-library(readr)
+# validate_data.R
 library(validate)
-simulated_data <- read_csv("data/00-simulated_data/simulated_data.csv")
+library(arrow)
+
+# Load simulated data
+simulated_data <- read_parquet("data/00-simulated_data/simulated_data.parquet")
+
+# Define validation rules
 rules <- validator(
-  !is.na(FWBscore) & FWBscore >= 1 & FWBscore <= 79,   # FWBscore: Check for non-null and range
-  !is.na(FSscore) & FSscore >= 1 & FSscore <= 77,       # FSscore: Check for non-null and range
-  !is.na(LMscore) & LMscore >= 0 & LMscore <= 3,         # LMscore: Check for non-null and range
-  !is.na(KHscore) & KHscore >= -3 & KHscore <= 1.267,    # KHscore: Check for non-null and range
-  !is.na(finalwt) & finalwt >= 0 & finalwt <= 6.64,   # finalwt: Check for non-null and range
-  !is.na(PUF_ID) & PUF_ID >= 7123 & PUF_ID <= 14400      # PUF_ID: Check for non-null and range
+  # Numeric ranges
+  !is.na(FWBscore) & FWBscore >= 1 & FWBscore <= 79,
+  !is.na(FSscore) & FSscore >= 1 & FSscore <= 77,
+  !is.na(LMscore) & LMscore >= 0 & LMscore <= 3,
+  !is.na(KHscore) & KHscore >= -2 & KHscore <= 1.267,
+  !is.na(PUF_ID) & PUF_ID >= 7123 & PUF_ID <= 14400,
+  
+  # Categorical variables validation
+  agecat %in% c("18-24", "25-34", "35-44", "45-54", "55-61", "62-69", "70-74", "75+"),
+  PPEDUC %in% c("Less than high school", "High school degree/GED", "Some college/Associate",
+                "Bachelor's degree", "Graduate/professional degree"),
+  PPINCIMP %in% c("Less than $20,000", "$20,000 to $29,999", "$30,000 to $39,999",
+                  "$40,000 to $49,999", "$50,000 to $59,999", "$60,000 to $74,999",
+                  "$75,000 to $99,999", "$100,000 to $149,999", "$150,000 or more"),
[…]

### 2024-12-03 scripts/03-test_analysis_data.R

@@ -8,18 +8,37 @@ library(readr)
 library(arrow)
 library(validate)
 analysis_data <- read_parquet("data/02-analysis_data/NFWBS_analysis_data.parquet")
+
+# Define validation rules
 rules <- validator(
-  !is.na(FWBscore) & FWBscore >= 1 & FWBscore <= 79,   # FWBscore: Check for non-null and range
-  !is.na(FSscore) & FSscore >= 1 & FSscore <= 77,       # FSscore: Check for non-null and range
-  !is.na(LMscore) & LMscore >= 0 & LMscore <= 3,         # LMscore: Check for non-null and range
-  !is.na(KHscore) & KHscore >= -3 & KHscore <= 1.267,    # KHscore: Check for non-null and range
-  !is.na(finalwt) & finalwt >= 0 & finalwt <= 6.64,   # finalwt: Check for non-null and range
-  !is.na(PUF_ID) & PUF_ID >= 7123 & PUF_ID <= 14400      # PUF_ID: Check for non-null and range
+  # Numeric ranges
+  !is.na(FWBscore) & FWBscore >= 1 & FWBscore <= 79,
+  !is.na(FSscore) & FSscore >= 1 & FSscore <= 77,
+  !is.na(LMscore) & LMscore >= 0 & LMscore <= 3,
+  !is.na(KHscore) & KHscore >= -2 & KHscore <= 1.267,
+  !is.na(PUF_ID) & PUF_ID >= 7123 & PUF_ID <= 14400,
+  
+  # Categorical variables validation
+  agecat %in% c("18-24", "25-34", "35-44", "45-54", "55-61", "62-69", "70-74", "75+"),
+  PPEDUC %in% c("Less than high school", "High school degree/GED", "Some college/Associate",
+                "Bachelor's degree", "Graduate/professional degree"),
+  PPINCIMP %in% c("Less than $20,000", "$20,000 to $29,999", "$30,000 to $39,999",
+                  "$40,000 to $49,999", "$50,000 to $59,999", "$60,000 to $74,999",
+                  "$75,000 to $99,999", "$100,000 to $149,999", "$150,000 or more"),
+  PPMARIT %in% c("Married", "Widowed", "Divorced/Separated", "Never married", "Living with partner"),
+  HEALTH %in% c("Poor", "Fair", "Good", "Very good", "Excellent"),
+  DISTRESS %in% c("Strongly disagree", "Disagree", "Neither agree nor disagree",
+                  "Agree", "Strongly agree"),
+  
+  # Structural checks
[…]

### 2024-12-03 paper/paper.qmd
Added cross references
Applied styler for the scripts

@@
 bibliography: references.bib
+toc: true
 ---
@@
 #| fig-width: 10
+#| label: fig-barplots
+#| fig-cap: "Bar Plots of Categorical Variables"
 
@@
 
-```{r}
-for (var in variables) {
-  print(var)
-  print(levels(data_cleaned[[var]]))
-}
-```
-
 # Model
@@
 All estimates are interpreted relative to these baseline categories."
+#| label: tbl-output
 
@@
 #| message: false
+#| label: tbl-suppcoef
 
@@
 Our analysis reveals several key patterns in the determinants of financial well-being.
-The coefficient plot above visualizes the main findings from our primary model, showing the relative importance of different factors.
+The coefficient table @tbl-output visualizes the main findings from our primary model, showing the relative importance of different factors.
 
@@
 ## Financial Knowledge and Skills
-Our supplementary model shows that all three measures of financial literacy contribute positively to financial well-being, though with varying magnitudes:
-- Financial skills (FSscore) show a significant positive effect (β = 0.501)
-- Knoll and Houts financial knowledge (KHscore) demonstrates the strongest effect (β = 3.748)
-- Lusardi and Mitchell financial knowledge (LMscore) shows a moderate positive effect (β = 0.874)
+Our supplementary model, as in @tbl-suppcoef, shows that all three measures of financial literacy contribute positively to financial well-being, though with varying magnitudes:
+- Financial skills (FSscore) show a significant positive effect ($\beta$ = 0.501)
+- Knoll and Houts financial knowledge (KHscore) demonstrates the strongest effect ($\beta$ = 3.748)
+- Lusardi and Mitchell financial knowledge (LMscore) shows a moderate positive effect ($\beta$ = 0.874)
 

### 2024-12-03 scripts/00-simulate_data.R

@@ -10,15 +10,21 @@ set.seed(2690) # Set seed for reproducibility
 
 # Define categorical variables and their levels
 agecat_levels <- c("18-24", "25-34", "35-44", "45-54", "55-61", "62-69", "70-74", "75+")
-ppeduc_levels <- c("Less than high school", "High school degree/GED", "Some college/Associate",
-                   "Bachelor's degree", "Graduate/professional degree")
-ppincimp_levels <- c("Less than $20,000", "$20,000 to $29,999", "$30,000 to $39,999",
-                     "$40,000 to $49,999", "$50,000 to $59,999", "$60,000 to $74,999",
-                     "$75,000 to $99,999", "$100,000 to $149,999", "$150,000 or more")
+ppeduc_levels <- c(
+  "Less than high school", "High school degree/GED", "Some college/Associate",
+  "Bachelor's degree", "Graduate/professional degree"
+)
+ppincimp_levels <- c(
+  "Less than $20,000", "$20,000 to $29,999", "$30,000 to $39,999",
+  "$40,000 to $49,999", "$50,000 to $59,999", "$60,000 to $74,999",
+  "$75,000 to $99,999", "$100,000 to $149,999", "$150,000 or more"
+)
 ppmarit_levels <- c("Married", "Widowed", "Divorced/Separated", "Never married", "Living with partner")
 health_levels <- c("Poor", "Fair", "Good", "Very good", "Excellent")
-distress_levels <- c("Strongly disagree", "Disagree", "Neither agree nor disagree",
-                     "Agree", "Strongly agree")
+distress_levels <- c(
+  "Strongly disagree", "Disagree", "Neither agree nor disagree",
+  "Agree", "Strongly agree"
+)
 
 # Number of observations
 n <- 1000

### 2024-12-03 scripts/02-test_simulated_data.R

@@ -19,19 +19,25 @@ rules <- validator(
   !is.na(LMscore) & LMscore >= 0 & LMscore <= 3,
   !is.na(KHscore) & KHscore >= -2 & KHscore <= 1.267,
   !is.na(PUF_ID) & PUF_ID >= 7123 & PUF_ID <= 14400,
-  
+
   # Categorical variables validation
   agecat %in% c("18-24", "25-34", "35-44", "45-54", "55-61", "62-69", "70-74", "75+"),
-  PPEDUC %in% c("Less than high school", "High school degree/GED", "Some college/Associate",
-                "Bachelor's degree", "Graduate/professional degree"),
-  PPINCIMP %in% c("Less than $20,000", "$20,000 to $29,999", "$30,000 to $39,999",
-                  "$40,000 to $49,999", "$50,000 to $59,999", "$60,000 to $74,999",
-                  "$75,000 to $99,999", "$100,000 to $149,999", "$150,000 or more"),
+  PPEDUC %in% c(
+    "Less than high school", "High school degree/GED", "Some college/Associate",
+    "Bachelor's degree", "Graduate/professional degree"
+  ),
+  PPINCIMP %in% c(
+    "Less than $20,000", "$20,000 to $29,999", "$30,000 to $39,999",
+    "$40,000 to $49,999", "$50,000 to $59,999", "$60,000 to $74,999",
+    "$75,000 to $99,999", "$100,000 to $149,999", "$150,000 or more"
+  ),
   PPMARIT %in% c("Married", "Widowed", "Divorced/Separated", "Never married", "Living with partner"),
   HEALTH %in% c("Poor", "Fair", "Good", "Very good", "Excellent"),
-  DISTRESS %in% c("Strongly disagree", "Disagree", "Neither agree nor disagree",
-                  "Agree", "Strongly agree"),
-  
+  DISTRESS %in% c(
+    "Strongly disagree", "Disagree", "Neither agree nor disagree",
+    "Agree", "Strongly agree"
+  ),
+
   # Structural checks
   is_unique(PUF_ID),
   nrow(.) == 1000,
@@ -43,4 +49,3 @@ validation_results <- confront(simulated_data, rules)
 # Generate validation report
 validation_summary <- summary(validation_results)
 print(validation_summary)
-

### 2024-12-03 scripts/03-test_analysis_data.R

@@ -17,19 +17,25 @@ rules <- validator(
   !is.na(LMscore) & LMscore >= 0 & LMscore <= 3,
   !is.na(KHscore) & KHscore >= -2 & KHscore <= 1.267,
   !is.na(PUF_ID) & PUF_ID >= 7123 & PUF_ID <= 14400,
-  
+
   # Categorical variables validation
   agecat %in% c("18-24", "25-34", "35-44", "45-54", "55-61", "62-69", "70-74", "75+"),
-  PPEDUC %in% c("Less than high school", "High school degree/GED", "Some college/Associate",
-                "Bachelor's degree", "Graduate/professional degree"),
-  PPINCIMP %in% c("Less than $20,000", "$20,000 to $29,999", "$30,000 to $39,999",
-                  "$40,000 to $49,999", "$50,000 to $59,999", "$60,000 to $74,999",
-                  "$75,000 to $99,999", "$100,000 to $149,999", "$150,000 or more"),
+  PPEDUC %in% c(
+    "Less than high school", "High school degree/GED", "Some college/Associate",
+    "Bachelor's degree", "Graduate/professional degree"
+  ),
+  PPINCIMP %in% c(
+    "Less than $20,000", "$20,000 to $29,999", "$30,000 to $39,999",
+    "$40,000 to $49,999", "$50,000 to $59,999", "$60,000 to $74,999",
+    "$75,000 to $99,999", "$100,000 to $149,999", "$150,000 or more"
+  ),
   PPMARIT %in% c("Married", "Widowed", "Divorced/Separated", "Never married", "Living with partner"),
   HEALTH %in% c("Poor", "Fair", "Good", "Very good", "Excellent"),
-  DISTRESS %in% c("Strongly disagree", "Disagree", "Neither agree nor disagree",
-                  "Agree", "Strongly agree"),
-  
+  DISTRESS %in% c(
+    "Strongly disagree", "Disagree", "Neither agree nor disagree",
+    "Agree", "Strongly agree"
+  ),
+
   # Structural checks
   is_unique(PUF_ID)
 )
@@ -40,5 +46,3 @@ validation_results <- confront(analysis_data, rules)
 # Generate validation report
 validation_summary <- summary(validation_results)
 print(validation_summary)
-
-

### 2024-12-04 other/datasheet/datasheet.qmd
Created datasheet
and added cross reference for sections

---
title: "Datasheet for CFPB Financial Well-Being Scale Dataset"
author: "Xinxiang Gao"
thanks: "Code and data are available at: [https://github.com/xgao28/financial_well_being_analysis](https://github.com/xgao28/financial_well_being_analysis)."
date: today
date-format: long
number-sections: true
bibliography: ../../paper/references.bib
format: pdf
---

# Datasheet for CFPB Financial Well-Being Scale Dataset

## Motivation

1. *For what purpose was the dataset created?*  
   The dataset was created to develop and validate a scale that measures individual financial well-being, aiming to provide a standardized tool for assessing this concept across diverse populations. It serves researchers, policymakers, and practitioners in understanding and improving financial health.

2. *Who created the dataset?*  
   The dataset was developed by the Consumer Financial Protection Bureau (CFPB) with contributions from experts in psychometrics and survey methodology. @cfpb_fwb_survey

## Composition

1. *What do the instances represent?*  
   Each instance represents an individual respondent's answers to the Financial Well-Being Scale items, along with demographic and other contextual information.

2. *How many instances are there in total?*  
   The dataset contains data from 6,394 respondents who participated in the national survey.

3. *Does the dataset contain all possible instances?*  
   The dataset includes a large representative sample of U.S. adults but does not cover every individual or every conceivable demographic subgroup.

4. *What data does each instance consist of?*  
   Each instance includes responses to the Financial Well-Being Scale, demographic characteristics (e.g., age, income), and potentially other financial behavior indicators.

5. *Is there a label or target associated with each instance?*  
   Yes, the primary target variable is the Financial Well-Being score derived from the scale items.

6. *Is any information missing from individual instances?*  
   Missing data were handled through rigorous preprocessing steps, ensuring completeness where possible.

7. *Are relationships between instances made explicit?*  
   Relationships between instances are not explicitly modeled within the dataset; each response is treated as independent.

8. *Are there recommended data splits?*  
   The dataset may be split into training and testing sets depending on the analysis objectives, though specific recommendations depend on the research question.

9. *Are there errors, noise, or redundancies?*  
   Efforts were made during data collection and preprocessing to minimize errors and noise, but some variability inherent to self-reported data may remain.

10. *Is the dataset self-contained?*  
    The dataset includes all necessary variables for calculating the Financial Well-Being Score, but external validation might require additional datasets.

11. *Does the dataset contain confidential data?*  
    No, the dataset anonymizes personal identifiers to protect respondent privacy.

12. *Does the dataset contain offensive or sensitive data?*  
    No, the content focuses on financial behaviors and attitudes without including sensitive personal information.

13. *Does the dataset identify subpopulations?*  
    Yes, it categorizes respondents by various demographic characteristics such as age, income level, and employment status.

14. *Is it possible to identify individuals?*  
    No, all personal identifiers have been removed to ensure anonymity.

15. *Does the dataset contain sensitive data?*  
    While it does not include directly sensitive attributes like race or religion, it does contain financial information which can be considered sensitive.

16. *Any other comments?*  
    None.

## Collection Process

1. *How was the data acquired?*  
   Data were collected through GfK's KnowledgePanel using address-based sampling methods.

2. *What mechanisms were used to collect the data?*  
   Mechanisms included online surveys and telephone interviews to accommodate different modes of participation.

3. *If the dataset is a sample, what was the sampling strategy?*  
   The sampling strategy aimed for a nationally representative sample, employing stratified random sampling techniques.

4. *Over what timeframe was the data collected?*  
   Data collection occurred over a specified period designed to capture seasonal variations in financial behavior.

5. *Were any ethical review processes conducted?*  
   Ethical considerations were addressed through IRB approval and adherence to best practices in human subjects research.

6. *Did you collect the data directly?*  
   The CFPB contracted with GfK to manage the data collection process.

7. *Were individuals notified about the data collection?*  
   Participants were informed about the study's purpose and provided consent prior to participation.

8. *Did individuals consent to data collection?*  
   Yes, all participants gave informed consent before taking part in the survey.

9. *Any other comments?*  
    None.

## Preprocessing, Cleaning, and Labeling

1. *Was any preprocessing done?*  
   Yes, extensive preprocessing was conducted, including cleaning, coding open-ended responses, and computing composite scores.

2. *Was raw data saved?*  
   Yes, raw data were preserved for future reference and verification purposes.

3. *Is the preprocessing software available?*  
   Preprocessing details are documented in the technical report accompanying the dataset.

4. *Any other comments?*  
   None.

## Uses

1. *Has the dataset been used for any tasks already?*  
   Yes, the dataset has been used to establish norms for the Financial Well-Being Scale and to conduct initial validity studies.

2. *What other tasks could the dataset be used for?*  
   The dataset can support research into financial education, policy evaluation, and intervention design aimed at improving financial well-being.

3. *Are there tasks for which the dataset should not be used?*  
   The dataset should not be used for identifying or targeting individuals due to anonymization measures.

4. *Any other comments?*  
   None.

## Maintenance

1. *Who will maintain the dataset?*  
   The CFPB is responsible for maintaining and updating the dataset as needed.

2. *How can the dataset owner be contacted?*  
   Contact information for inquiries regarding the dataset can be found on the CFPB website.

3. *Will the dataset be updated?*  
   Updates may occur periodically to reflect changes in financial circumstances and new research findings.

4. *Any other comments?*  
   None.

# References

### 2024-12-04 paper/paper.qmd
Created datasheet
and added cross reference for sections

@@
 
-Our findings indicate [brief summary of key results], suggesting important implications for financial education programs and policy interventions.
-These results highlight the need for tailored approaches that consider both financial knowledge and individual circumstances in promoting financial well-being.
+Our findings indicate that psychological distress has the strongest negative association with financial well-being (-16.5 points), while higher income ($150,000+) shows the largest positive effect (+12.6 points).
+Advanced age (75+ years) and excellent health status are also strong positive predictors (+9.4 and +7.7 points respectively).
+Notably, financial knowledge measures show relatively modest effects compared to these psychological and demographic factors, suggesting a need to rethink traditional approaches to financial education.
 
-The rest of the paper is organized as follows: Section 2 reviews relevant literature on financial well-being and its determinants.
-Section 3 describes our data and methodology.
-Section 4 presents our empirical results.
-Section 5 discusses the implications of our findings and concludes.
-The appendix provides additional details on the survey methodology and sampling design.
+The rest of the paper is organized as follows: @sec-data describes our data and methodology.
+@sec-model presents our modeling approach.
+@sec-results presents our empirical results.
+@sec-discussion discusses the implications of our findings and concludes.
+@sec-app provides additional details on the survey methodology and @sec-ack provides acknowledgements.
 
-# Data
+# Data {#sec-data}
 
@@
 
-The dataset for this study is the Consumer Financial Protection Bureau’s (CFPB) National Financial Well-Being Survey Public Use File (PUF), collected in 2017.
+The dataset for this study is the National Financial Well-Being Survey Public Use File (PUF) collected by @cfpb_fwb_survey.
 This dataset is a rich resource that includes 217 variables encompassing financial, demographic, and psychosocial factors, drawn from a representative sample of the U.S.
@@
 
+## Variable Summary
+
+```{r}
+#| echo: false
+#| warning: false
+#| message: false
+#| label: fig-histograms
+#| fig-cap: "Histograms of Financial Scores"
+data_long <- data_cleaned %>%
+  pivot_longer(
+    cols = c(FWBscore, FSscore, LMscore, KHscore),
+    names_to = "ScoreType", values_to = "Value"
+  )
+
+ggplot(data_long, aes(x = Value)) +
+  geom_histogram(binwidth = 1, color = "white", fill = "steelblue", alpha = 0.9) +
+  facet_wrap(~ScoreType, scales = "free", ncol = 2) +
+  labs(
+    # title = "Histograms of Financial Scores",
+    # subtitle = "Distribution of scores across different financial measures",
+    # caption = "FWBscore: Financial Well-being Scale\nFSscore: Financial Skill Scale\nLMscore: Lusardi and Mitchell Financial Knowledge Scale\nKHscore: Knoll and Houts Financial Knowledge Scale",
+    x = "Score Value",
+    y = "Frequency"
+  ) +
+  theme_minimal(base_size = 14) +
+  theme(
+    plot.title = element_text(face = "bold", hjust = 0.5),
+    plot.subtitle = element_text(hjust = 0.5),
+    strip.background = element_rect(fill = "lightgray"),
+    strip.text = element_text(face = "bold")
+  )
+```
+
+@fig-histograms shows the distribution of four different financial scores (FSscore, FWBscore, KHscore, and LMscore) through histograms.
+FSscore and FWBscore show continuous distributions with bell-shaped curves, though not perfectly normal: FSscore ranges from about 0-80 with a peak around 40-50, while FWBscore has a similar range but appears more symmetrically distributed with its peak around 45-50.
+The KHscore shows a discrete distribution across four main categories (from -2 to 1), with the highest frequency at 0 and relatively balanced frequencies between -1 and 1.
+The LMscore is also discrete but shows a strong positive skew, with values ranging from 0 to 3 and a clear majority of responses at the highest value (3).
+This pattern suggests that while financial wellbeing and financial situations (FWB and FS) vary normally across the population, the literacy measure (LM) indicates most respondents have high literacy levels.
+
 ```{r}
@@
 ```
+
+The correlation matrix heatmap as @fig-correlations shows the relationships between four different scores: FWBscore, FSscore, LMscore, and KHscore.
+The strongest correlation appears between LMscore and KHscore with a moderate positive correlation of 0.6, followed by a moderate correlation of 0.49 between FWBscore and FSscore.
+FWBscore shows weak to moderate positive correlations with LMscore (0.23) and KHscore (0.33), while FSscore exhibits the weakest correlations with LMscore (0.11) and KHscore (0.19).
+All correlations are positive, suggesting that as one score increases, the others tend to increase as well, though with varying strengths of association.
+The diagonal shows perfect correlations of 1.0, as expected when comparing a variable with itself.
 
@@
 
-```{r}
-#| echo: false
-#| warning: false
-#| message: false
-data_long <- data_cleaned %>%
-  pivot_longer(
-    cols = c(FWBscore, FSscore, LMscore, KHscore),
-    names_to = "ScoreType", values_to = "Value"
-  )
+@fig-barplots provides a comprehensive demographic overview of the survey population.
+The age distribution shows strong representation across middle-age groups (25-34, 45-54, and 62-69) with fewer respondents in the youngest (18-24) and oldest (70-74) categories.
+Education levels indicate that "Some college/Associate" and "High school degree/GED" are the most common, while "Less than high school" is the least common.
+Income distribution is relatively balanced across categories, with slightly higher numbers in the $100,000-$149,999 range.
+Marital status shows a clear majority of "Married" respondents, significantly outnumbering other categories.
+Health status data reveals most respondents report "Very good" or "Good" health, with few reporting "Poor" health.
+Finally, psychological distress measurements show a normal-like distribution centered around "Neither agree nor disagree" and "Agree" responses, with fewer responses at the extremes ("Strongly agree" and "Strongly disagree").
 
-ggplot(data_long, aes(x = Value)) +
-  geom_histogram(binwidth = 1, color = "white", fill = "steelblue", alpha = 0.9) +
-  facet_wrap(~ScoreType, scales = "free", ncol = 2) +
-  labs(
-    title = "Histograms of Financial Scores",
-    subtitle = "Distribution of scores across different financial measures",
-    # caption = "FWBscore: Financial Well-being Scale\nFSscore: Financial Skill Scale\nLMscore: Lusardi and Mitchell Financial Knowledge Scale\nKHscore: Knoll and Houts Financial Knowledge Scale",
-    x = "Score Value",
-    y = "Frequency"
-  ) +
-  theme_minimal(base_size = 14) +
-  theme(
-    plot.title = element_text(face = "bold", hjust = 0.5),
-    plot.subtitle = element_text(hjust = 0.5),
-    strip.background = element_rect(fill = "lightgray"),
-    strip.text = element_text(face = "bold")
-  )
-```
-
-# Model
+# Model {#sec-model}
 
@@
 Our primary model examines the relationship between financial well-being and demographic and psychosocial factors, while a supplementary model focuses on the role of financial knowledge and skills.
+
+## Primary Model
 
@@
 
+## Supplementary Model
+
 The supplementary model examining financial literacy's role is:
@@
 #| fig-cap: "Model Diagnostic Plots"
+#| fig-align: 'center'
+#| fig-width: 10
 #| fig-height: 8
-#| fig-width: 10
-#| layout: [[1,1], [1,1]]
-#| echo: false
 #| warning: false
 #| message: false
-# Preprocess data: Ensure predictors are factors and drop missing values
+#| echo: false
+
+library(tidyverse)
+library(gridExtra)
+
+# Preprocess data and fit model
 data_model <- data_cleaned %>%
-  mutate(across(all_of(c("agecat", "PPEDUC", "PPINCIMP", "PPMARIT", "HEALTH", "DISTRESS")), as.factor))
+  mutate(across(all_of(c("agecat", "PPEDUC", "PPINCIMP", "PPMARIT", 
+                        "HEALTH", "DISTRESS")), as.factor))
 
-# Fit the linear regression model
-lm_model <- lm(FWBscore ~ agecat + PPEDUC + PPINCIMP + PPMARIT + HEALTH + DISTRESS, data = data_model)
+# Fit the model
+lm_model <- lm(FWBscore ~ agecat + PPEDUC + PPINCIMP + PPMARIT + 
+               HEALTH + DISTRESS, data = data_model)
 
-par(mfrow = c(2, 2))
-plot(lm_model, which = 1:4)
+# Create the four diagnostic plots
+# Residuals vs Fitted
+p1 <- ggplot(data.frame(
+    fitted = fitted(lm_model),
+    residuals = resid(lm_model)
+  ), aes(x = fitted, y = residuals)) +
+  geom_point(alpha = 0.3) +
+  geom_smooth(method = "loess", se = FALSE, color = "red") +
+  geom_hline(yintercept = 0, linetype = "dashed") +
+  labs(title = "Residuals vs Fitted",
+       x = "Fitted values",
+       y = "Residuals") +
+  theme_minimal()
+
+# Normal Q-Q
+p2 <- ggplot(data.frame(
+    theoretical = qnorm(ppoints(length(resid(lm_model)))),
+    sample = sort(scale(resid(lm_model)))
+  ), aes(x = theoretical, y = sample)) +
+  geom_point(alpha = 0.3) +
+  geom_abline(slope = 1, intercept = 0, color = "red") +
+  labs(title = "Normal Q-Q",
+       x = "Theoretical Quantiles",
+       y = "Standardized Residuals") +
+  theme_minimal()
+
+# Scale-Location
+p3 <- ggplot(data.frame(
+    fitted = fitted(lm_model),
+    sqrt_abs_resid = sqrt(abs(scale(resid(lm_model))))
+  ), aes(x = fitted, y = sqrt_abs_resid)) +
+  geom_point(alpha = 0.3) +
+  geom_smooth(method = "loess", se = FALSE, color = "red") +
+  labs(title = "Scale-Location",
+       x = "Fitted values",
+       y = expression(sqrt("|Standardized residuals|"))) +
+  theme_minimal()
+
+# Cook's Distance
+p4 <- ggplot(data.frame(
+    index = 1:length(cooks.distance(lm_model)),
+    cooks = cooks.distance(lm_model)
+  ), aes(x = index, y = cooks)) +
+  geom_point(alpha = 0.3) +
+  labs(title = "Cook's Distance",
+       x = "Observation Number",
+       y = "Cook's distance") +
+  theme_minimal()
+
+# Combine plots
+grid.arrange(p1, p2, p3, p4, ncol = 2)
 ```
 
-We conducted several diagnostic checks to validate our model assumptions.
+We conducted several diagnostic checks to validate our model assumptions, as illustrated in @fig-model-diagnostics.
 The residual plots show relatively homoscedastic errors and no serious violations of linearity.
@@
 
-# Results
+# Results {#sec-results}
 
@@
 lm_results <- data.frame(
-  Term = rownames(model_summary$coefficients),
-  Estimate = model_summary$coefficients[, "Estimate"],
-  Std_Error = model_summary$coefficients[, "Std. Error"],
-  t_value = model_summary$coefficients[, "t value"],
-  p_value = model_summary$coefficients[, "Pr(>|t|)"]
+  # Term = rownames(model_summary$coefficients),
+  estimate = model_summary$coefficients[, "Estimate"],
+  std.error = model_summary$coefficients[, "Std. Error"],
+  statistic = model_summary$coefficients[, "t value"],
+  p.value = model_summary$coefficients[, "Pr(>|t|)"]
 )
@@
   lm_results,
-  caption = "Summary of Linear Regression Model"
+  caption = "Summary of Supplementary Model"
 )
@@
 ## Financial Knowledge and Skills
-Our supplementary model, as in @tbl-suppcoef, shows that all three measures of financial literacy contribute positively to financial well-being, though with varying magnitudes:
-- Financial skills (FSscore) show a significant positive effect ($\beta$ = 0.501)
-- Knoll and Houts financial knowledge (KHscore) demonstrates the strongest effect ($\beta$ = 3.748)
-- Lusardi and Mitchell financial knowledge (LMscore) shows a moderate positive effect ($\beta$ = 0.874)
+Our supplementary model, as in Table 2, requires careful interpretation given the distinct distributions of our financial literacy measures:
 
-These results suggest that while financial knowledge and skills are important, demographic and psychosocial factors play substantial roles in determining financial well-being.
-The particularly strong negative effect of psychological distress highlights the importance of mental health in financial outcomes.
+Financial skills (FSscore) shows a moderate positive effect ($\beta$ = 0.501), which is meaningful given its approximately normal distribution across the population.
 
-# Discussion
+While Knoll and Houts financial knowledge (KHscore) shows a large coefficient ($\beta$ = 3.748), this should be interpreted with caution given its discrete, roughly uniform distribution across four categories (-2 to 1).
+
+The Lusardi and Mitchell financial knowledge (LMscore) effect ($\beta$ = 0.874) must be interpreted carefully due to its highly skewed distribution, with most respondents scoring at the maximum value of 3.
+
+These results suggest that while all measures show positive associations with financial well-being, their interpretations need to account for their distinct distributional characteristics.
+The FSscore's effect may be the most reliably interpretable given its normal distribution, while the strong coefficient for KHscore and moderate coefficient for LMscore should be considered in light of their non-normal distributions.
+These distributional patterns suggest that future research might benefit from more granular measures of financial knowledge that could better capture variability across the population.
+
+# Discussion {#sec-discussion}
 
@@
 
-## Appendix: Survey Methodology and Sampling Design
+## Survey Methodology {#sec-app}
 
-### Survey Design and Implementation
+The CFPB Financial Well-Being Scale methodology merits critical examination across three key dimensions: sampling design, measurement validation, and methodological limitations.
 
-#### Sampling Framework
-The CFPB National Financial Well-Being Survey employs a sophisticated sampling design through GfK's KnowledgePanel.
-The panel uses address-based sampling (ABS) methods derived from the United States Postal Service's Delivery Sequence File.
-This approach offers several advantages over traditional random-digit dialing:
+### Sampling Framework and Implementation
 
-1. **Coverage**: The sampling frame covers approximately 97% of U.S. households, including those without landline phones or internet access.
+The survey employed a multi-round data collection approach through a large nationally recognized provider of survey data.
+The initial round fielded 46 candidate items to 4,500 respondents, with the sample divided between adults aged 18 to 61 and those aged 62 and older.
+This approach aimed for a diverse consumer base, though not necessarily nationally representative, ensuring that the sample was roughly similar to the U.S.
+population based on Census demographics for working-age and older consumers.
+The inclusion of both younger and older age groups, as well as varying modes of administration (self-administered and interviewer-administered), is a strength in addressing potential mode effects and providing insights into how different populations respond to the scale.
 
-2. **Non-Response Mitigation**: By providing internet access and devices to selected households that lack them, the design reduces the selection bias typically associated with online surveys.
+However, despite efforts to achieve diversity, there are inherent challenges in fully representing transient or hard-to-reach populations, which could introduce selection bias.
+Moreover, the exclusion of approximately 500 pencil-and-paper surveys due to lack of confidence in these data suggests variability in data quality across different collection methods, potentially affecting the generalizability of findings.
 
-3. **Panel Characteristics**: The ongoing nature of the panel allows for:
-   - Detailed profiling of panelists
-   - Lower attrition rates compared to one-time surveys
-   - Higher response rates due to established relationships
+### Measurement Development and Validation
 
-#### Sample Size and Power Analysis
-The final sample of 6,394 respondents was determined through power analysis to detect small-to-medium effect sizes (Cohen's d = 0.2) with 80% power at α = 0.05. This sample size accounts for:
+The development of the Financial Well-Being Scale adhered to item response theory (IRT) principles, involving extensive rounds of refinement from an initial pool of items down to the final set.
+The use of exploratory and confirmatory factor analysis, alongside IRT modeling, provided robust evidence for the scale's structure.
+Psychometric analyses conducted by Vector Psychometric Group supported the use of a bi-factor model, capturing an overall financial well-being score while accounting for positive and negative wording effects within items.
 
-- Expected response rates (~64%)
-- Need for subgroup analyses
-- Anticipated missing data
-- Design effects from complex sampling
+The reported marginal reliability of at least 0.89 indicates strong internal consistency, and the abbreviated version also showed acceptable reliability above 0.80.
+However, the absence of temporal stability metrics limits the understanding of the scale's consistency over time.
+Furthermore, the reliance on cognitive interviews with a relatively small sample size (n=19) may limit the breadth of perspectives captured, especially across diverse populations.
 
-#### Survey Administration
-The survey implementation incorporated several methodological features to enhance data quality:
+Validation of the scale included testing its performance across different demographic groups and survey modes, supporting the comparability of scores derived from different contexts.
+Known-groups validation helped establish discriminant validity, and correlations with objective financial measures provided criterion validity.
+Nevertheless, cross-validation with external datasets remains limited, which can affect the extent to which the results can be generalized beyond the study sample.
 
-1. **Mixed-Mode Testing**:
-   - Self-administered online version
-   - Interviewer-administered telephone version
-   - Comparative analysis of mode effects
+### Methodological Strengths and Limitations
 
-2. **Response Rate Enhancement**:
-   - Multiple contact attempts
-   - Incentive structure for participation
-   - Reminder emails and phone calls
+The mixed-mode implementation allowed for the assessment of mode effects, which is important given documented differences in self-reporting behaviors depending on the survey format.
+Quality control measures such as attention checks and monitoring of completion times were used to ensure data integrity, although the specific criteria applied and their impact on the final sample composition are not detailed.
 
-3. **Quality Control Measures**:
-   - Attention check questions
-   - Time-to-completion monitoring
-   - Response pattern analysis
+Critical limitations include:
+1. Potential underrepresentation of certain populations despite the address-based sampling methodology.
+2. Self-report bias in financial measures, particularly relevant for sensitive topics.
+3. Limited assessment of the temporal stability of the financial well-being construct.
 
-### Measurement Validation
+The study achieved a response rate that exceeds typical online panel rates, but this also raises questions about nonresponse bias.
+While multiple contact attempts and incentives were utilized, a deeper analysis of nonresponse patterns would enhance methodological transparency.
 
-#### Scale Development Process
-The Financial Well-Being Scale underwent rigorous development:
+In summary, the methodological choices reflect a balance between practical constraints and ideal survey design standards.
+The approach aligns with contemporary practices in survey research while acknowledging limitations common in large-scale financial studies.
 
-1. **Initial Item Pool**:
-   - 47 candidate items
-   - Cognitive interviews with 19 adults
-   - Expert review panel assessment
+\newpage
 
-2. **Iterative Testing**:
-   - Three rounds of data collection
-   - Progressive item reduction
-   - Final selection of 10 items
-
-3. **Psychometric Analysis**:
-   - Item Response Theory modeling
-   - Differential Item Functioning analysis
-   - Reliability assessment (marginal reliability ≥ 0.89)
-
-#### Cross-Validation
-The scale's validity was established through multiple approaches:
-
-1. **Content Validity**:
-   - Expert review
-   - Consumer feedback
-   - Theoretical alignment
-
-2. **Construct Validity**:
-   - Factor analysis
-   - Known-groups validation
-   - Convergent/discriminant validity
-
-3. **Criterion Validity**:
-   - Correlation with objective financial measures
-   - Predictive validity assessment
-   - Cross-validation with external datasets
-
-### Limitations and Considerations
-
-1. **Selection Bias**:
-   - While the KnowledgePanel mitigates many selection issues, some hard-to-reach populations may still be underrepresented
-   - Panel conditioning effects may influence responses
-
-2. **Measurement Error**:
-   - Self-reported data subject to social desirability bias
-   - Recall bias in retrospective questions
-   - Mode effects between self-administered and interviewer-administered versions
-
-3. **External Validity**:
-   - Generalizability across different economic conditions
-   - Temporal stability of constructs
-   - Cross-cultural applicability
-
-### Simulation Study
-To assess the robustness of our sampling approach, we conducted a simulation study:
-
-1. **Method**:
-   - Generated 1,000 synthetic datasets
-   - Varied response patterns and missing data mechanisms
-   - Tested sensitivity to different sampling weights
-
-2. **Results**:
-   - Sample size provides stable estimates (CV < 0.05)
-   - Minimal bias from missing data (< 2% change in estimates)
-   - Robust to different weighting schemes
-
-3. **Implications**:
-   - Confirms adequacy of sample size
-   - Supports reliability of estimates
-   - Validates analytical approach
-
-This methodological framework ensures high-quality data collection while acknowledging and addressing potential limitations, providing a solid foundation for analyzing financial well-being across diverse populations.
-
-## Acknowledgements
+## Acknowledgements {#sec-ack}
 

### 2024-12-04 paper/paper.qmd
Updated sketches and added explanation
of variable selection in the paper.

@@
 Stress (DISTRESS): Perceived level of stress, rated on a 5-point Likert scale from 1 (Not at all) to 5 (Extremely).
+
+Note that this study employs categorical grouping for variables that could be continuous (such as age and income).
+This approach offers several advantages in survey research: it improves interpretability of results across demographic groups, facilitates statistical comparisons between meaningful socioeconomic segments, and aligns with typical policy and program planning categories.
+For sensitive variables like income and age, categorical grouping also helps protect respondent privacy while maintaining sufficient detail for analysis.
+Additionally, the use of standardized categories (in age, education, and income) enables easier comparison with other survey research and demographic studies.
 

### 2024-12-04 paper/paper.qmd
Update paper for prose concerns

--- 2024-12-04 paper/paper.qmd
@@
 ---
-title: "Determinants of Financial Well-Being: The Role of Demographics, Psychosocial Factors, and Financial Literacy"
-subtitle: "Psychological Distress Has the Strongest Negative Impact on Financial Outcomes"
+title: "Factors Shaping Financial Well-Being: Examining Demographics, Psychological Elements, and Financial Knowledge"
+subtitle: "Psychological Distress and Financial Health: Measuring the Relationship"
 author: "Xinxiang Gao"
@@
 date-format: long
-abstract: "This paper analyzes the determinants of financial well-being using the Consumer Financial Protection Bureau's National Financial Well-Being Survey.
-Using a comprehensive regression analysis of over 6,000 respondents, we find that psychological distress has the strongest negative association with financial well-being, with highly stressed individuals scoring 16.5 points lower on the financial well-being scale.
-Higher income and age are associated with better financial outcomes, while education shows modest positive effects.
-These findings suggest that financial education programs should incorporate mental health support, particularly for vulnerable populations."
+abstract: "This paper examines factors affecting financial well-being using the Consumer Financial Protection Bureau's National Financial Well-Being Survey.
+Through regression analysis of over 6,000 respondents, we found that psychological distress showed the strongest negative connection to financial well-being, with highly stressed individuals scoring 16.5 points lower on the financial well-being scale.
+People with higher income and those in older age groups reported better financial outcomes, while education demonstrated small positive effects.
+These results indicate that financial education programs would benefit from including mental health support, especially for people facing financial challenges."
 format: pdf
@@
 # Introduction
+Financial well-being forms a core objective for individuals and households navigating modern economic conditions.
+The concept encompasses meeting present and future financial obligations, influenced by financial literacy, demographics, and psychological factors.
+Understanding these relationships informs policy design and programs to enhance financial outcomes across populations.
 
-In today's complex financial landscape, achieving financial well-being is a fundamental goal for individuals and households.
-While financial well-being encompasses the ability to meet current and future financial obligations, its determinants span multiple domains including financial literacy, demographic characteristics, and psychosocial factors.
-Understanding these relationships is crucial for developing effective policies and interventions to improve financial outcomes across diverse population segments.
+The Consumer Financial Protection Bureau (CFPB) created the Financial Well-Being Scale (FWBscore), a 10-item questionnaire measuring four components: control over daily finances, ability to handle financial emergencies, progress on financial objectives, and freedom to make life choices.
+Researchers have developed several tools to evaluate financial knowledge.
+The Lusardi and Mitchell financial knowledge scale (LMscore) by @lusardi2008 uses three items to measure basic financial concepts.
+The Knoll and Houts financial knowledge scale (KHscore) by @knoll2012 employs a 10-item assessment based on item response theory.
+The CFPB Financial Skill Scale (FSscore) by @cfpb2015 measures practical financial abilities using IRT methods.
 
-To measure financial well-being systematically, the Consumer Financial Protection Bureau (CFPB) developed the Financial Well-Being Scale (FWBscore), a 10-item questionnaire assessing four critical elements: control over day-to-day finances, capacity to absorb financial shocks, progress toward financial goals, and ability to make life-improving choices.
-Alongside this measure, researchers have developed various instruments to assess financial knowledge and skills.
-The Lusardi and Mitchell financial knowledge scale (LMscore) by @lusardi2008 provides a foundational three-item measure of basic financial concepts, while the Knoll and Houts financial knowledge scale (KHscore) by @knoll2012 offers a more comprehensive 10-item assessment using item response theory.
-Additionally, the CFPB Financial Skill Scale (FSscore) by @cfpb2015 evaluates practical financial competencies through an IRT-based approach.
+Research has studied how financial literacy affects economic outcomes.
+However, questions remain about how demographics and psychological factors combine with financial knowledge to shape financial well-being.
+Understanding these relationships could better explain differences in financial outcomes across population groups.
 
-While previous research has examined financial literacy's role in economic outcomes, less attention has been paid to how demographic and psychosocial factors interact with financial knowledge to influence overall financial well-being.
-This gap is particularly significant given evidence suggesting that financial knowledge alone may not fully explain variations in financial outcomes across different population segments.
+Our study examines how changes in demographic and psychological characteristics relate to financial well-being scores, while controlling for other variables.
+This analysis identifies which factors most strongly influence financial well-being, helping guide policy decisions.
 
-Our primary estimand is the average change in financial well-being score associated with variations in demographic and psychosocial characteristics, holding other factors constant.
-This causal parameter helps identify which factors have the strongest relationship with financial well-being, enabling targeted policy interventions.
+The results show psychological distress has the strongest negative link to financial well-being (-16.5 points).
+Higher income ($150,000+) shows the largest positive relationship (+12.6 points).
+Age over 75 years and excellent health status also predict better outcomes (+9.4 and +7.7 points respectively).
+Financial knowledge measures show smaller effects than these psychological and demographic factors, suggesting new approaches to financial education may be needed.
 
-Our findings indicate that psychological distress has the strongest negative association with financial well-being (-16.5 points), while higher income ($150,000+) shows the largest positive effect (+12.6 points).
-Advanced age (75+ years) and excellent health status are also strong positive predictors (+9.4 and +7.7 points respectively).
-Notably, financial knowledge measures show relatively modest effects compared to these psychological and demographic factors, suggesting a need to rethink traditional approaches to financial education.
-
-The rest of the paper is organized as follows: @sec-data describes our data and methodology.
-@sec-model presents our modeling approach.
-@sec-results presents our empirical results.
-@sec-discussion discusses the implications of our findings and concludes.
-@sec-app provides additional details on the survey methodology and @sec-ack provides acknowledgements.
+The paper continues as follows: @sec-data describes our data and methods.
+@sec-model explains our modeling approach.
+@sec-results presents our findings.
+@sec-discussion examines implications and concludes.
+@sec-app details survey methods and @sec-ack lists acknowledgements.
 
@@
 
-The dataset for this study is the National Financial Well-Being Survey Public Use File (PUF) collected by @cfpb_fwb_survey.
-This dataset is a rich resource that includes 217 variables encompassing financial, demographic, and psychosocial factors, drawn from a representative sample of the U.S.
-population.
-It provides data on key indicators such as financial knowledge, skills, behaviors, and overall well-being.
-These data are ideal for analyzing relationships between financial well-being and other financial or non-financial factors.
+The National Financial Well-Being Survey Public Use File (PUF) by @cfpb_fwb_survey provides the data for this study.
+The dataset contains 217 variables measuring financial, demographic, and psychological factors from a representative U.S.
+population sample.
+It records financial knowledge, skills, behaviors, and well-being.
 
-The survey was conducted using the GfK KnowledgePanel, an online probability-based panel designed to be representative of the U.S.
-adult population.
-The GfK KnowledgePanel is created by randomly selecting households and inviting them to participate in the panel.
-Households without internet access are provided with a computer and internet service.
-This method ensures that the sample is representative across various demographic groups in the United States.
-Additionally, the dataset incorporates external contextual variables, such as poverty levels in respondents’ counties of residence, offering a nuanced perspective on financial well-being in diverse settings.
-While other datasets, such as the Survey of Consumer Finances or the Panel Study of Income Dynamics, also include financial information, they do not provide a standardized, validated measure of financial well-being like the FWBscore.
-This makes the CFPB dataset uniquely suited for this study.
+GfK KnowledgePanel collected the survey data using an online probability-based panel representing U.S.
+adults.
+The panel forms through random household selection and invitation.
+GfK provides computers and internet service to households lacking access, ensuring representation across U.S.
+demographic groups.
+The dataset includes contextual variables like county-level poverty rates, providing environmental context for financial well-being assessments.
+While other datasets like the Survey of Consumer Finances and Panel Study of Income Dynamics contain financial information, they lack the standardized FWBscore measure, making this CFPB dataset particularly suitable.
 
@@
 ### Primary Outcome Variable
-
-Financial Well-Being Score (FWBscore): The CFPB Financial Well-Being Scale score is the key outcome variable.
-It is a 10-item scale capturing four dimensions: control over day-to-day finances, capacity to absorb financial shocks, progress toward financial goals, and freedom of financial choices.
-Scores range from 0 to 100, with higher scores indicating greater financial well-being.
+Financial Well-Being Score (FWBscore): This 10-item CFPB scale measures four areas: control over day-to-day finances, capacity to handle financial emergencies, progress toward financial goals, and freedom of financial choices.
+Scores range from 0 to 100, with higher scores showing better financial well-being.
 
@@
 
-All the independent variables are collected through self-reported questionnaire responses, as addressed below.
+The data comes from self-reported questionnaire responses:
 
-Age (agecat): Categorized into age groups (e.g., 18–24, 25–34, 35–44, 45–54, 55–64, 65+).
+Age (agecat): Groups include 18–24, 25–34, 35–44, 45–54, 55–64, and 65+.
 
-Education (PPEDUC): Highest degree attained, categorized from less than high school to graduate degrees.
+Education (PPEDUC): Categories range from less than high school to graduate degrees.
 
-Household Income (PPINCIMP): Categorized income levels (e.g., <$20,000, $20,000-$39,999, $40,000-$59,999, etc.).
+Household Income (PPINCIMP): Groups span from <$20,000 to higher brackets like $20,000-$39,999 and $40,000-$59,999.
 
-Marital Status (PPMARIT): Categories include single, married, divorced, widowed, and other.
+Marital Status (PPMARIT): Includes single, married, divorced, widowed, and other.
 
-Health (HEALTH): Self-reported general health, rated on a 5-point Likert scale from 1 (Poor) to 5 (Excellent).
+Health (HEALTH): Self-reported rating from 1 (Poor) to 5 (Excellent).
 
-Stress (DISTRESS): Perceived level of stress, rated on a 5-point Likert scale from 1 (Not at all) to 5 (Extremely).
+Stress (DISTRESS): Self-reported rating from 1 (Not at all) to 5 (Extremely).
+
+Using categories for continuous variables like age and income helps interpret results across demographic groups, enables statistical group comparisons, and matches policy planning needs.
+This grouping also protects respondent privacy while maintaining analytical detail.
+The standard categories allow comparison with other demographic research.
 
@@
 
-Lusardi and Mitchell Financial Knowledge Scale Score (LMscore): A 3-item summative scale measuring financial literacy, particularly in planning and decision-making.
+Lusardi and Mitchell Financial Knowledge Scale Score (LMscore): Three items measuring financial literacy in planning and decision-making.
 
-Knoll and Houts Financial Knowledge Scale Score (KHscore): A 10-item IRT-based measure assessing broader financial knowledge across multiple dimensions.
+Knoll and Houts Financial Knowledge Scale Score (KHscore): Ten items using IRT methods to assess financial knowledge.
 
-Financial Skill Scale Score (FSscore): An IRT-based measure developed by the CFPB, capturing practical financial skills like budgeting, debt management, and saving.
+Financial Skill Scale Score (FSscore): CFPB-developed IRT measure of practical financial skills including budgeting, debt management, and saving.
 
@@
 
-@fig-histograms shows the distribution of four different financial scores (FSscore, FWBscore, KHscore, and LMscore) through histograms.
-FSscore and FWBscore show continuous distributions with bell-shaped curves, though not perfectly normal: FSscore ranges from about 0-80 with a peak around 40-50, while FWBscore has a similar range but appears more symmetrically distributed with its peak around 45-50.
-The KHscore shows a discrete distribution across four main categories (from -2 to 1), with the highest frequency at 0 and relatively balanced frequencies between -1 and 1.
-The LMscore is also discrete but shows a strong positive skew, with values ranging from 0 to 3 and a clear majority of responses at the highest value (3).
-This pattern suggests that while financial wellbeing and financial situations (FWB and FS) vary normally across the population, the literacy measure (LM) indicates most respondents have high literacy levels.
+@fig-histograms displays distributions of four financial measures: FSscore, FWBscore, KHscore, and LMscore.
+The FSscore and FWBscore follow roughly bell-shaped patterns.
+FSscore spans 0-80, peaking between 40-50, while FWBscore shows a more balanced distribution centered at 45-50.
+KHscore plots across four points (-2 to 1), with the highest frequency at 0 and similar frequencies at -1 and 1.
+LMscore ranges from 0 to 3, with most responses concentrated at 3.
+These patterns indicate that financial well-being and financial situations distribute evenly through the population, while most respondents score high on basic financial literacy.
 
@@
 
-The correlation matrix heatmap as @fig-correlations shows the relationships between four different scores: FWBscore, FSscore, LMscore, and KHscore.
-The strongest correlation appears between LMscore and KHscore with a moderate positive correlation of 0.6, followed by a moderate correlation of 0.49 between FWBscore and FSscore.
-FWBscore shows weak to moderate positive correlations with LMscore (0.23) and KHscore (0.33), while FSscore exhibits the weakest correlations with LMscore (0.11) and KHscore (0.19).
-All correlations are positive, suggesting that as one score increases, the others tend to increase as well, though with varying strengths of association.
-The diagonal shows perfect correlations of 1.0, as expected when comparing a variable with itself.
+@fig-correlations presents a correlation matrix heatmap comparing FWBscore, FSscore, LMscore, and KHscore.
+LMscore and KHscore show the strongest relationship (0.6), followed by FWBscore and FSscore (0.49).
+FWBscore pairs with LMscore (0.23) and KHscore (0.33), showing smaller positive relationships.
+FSscore has the smallest relationships with LMscore (0.11) and KHscore (0.19).
+The positive values indicate that scores tend to move in the same direction.
+Each measure's correlation with itself equals 1.0 on the diagonal.
 
@@
 
-@fig-barplots provides a comprehensive demographic overview of the survey population.
-The age distribution shows strong representation across middle-age groups (25-34, 45-54, and 62-69) with fewer respondents in the youngest (18-24) and oldest (70-74) categories.
-Education levels indicate that "Some college/Associate" and "High school degree/GED" are the most common, while "Less than high school" is the least common.
-Income distribution is relatively balanced across categories, with slightly higher numbers in the $100,000-$149,999 range.
-Marital status shows a clear majority of "Married" respondents, significantly outnumbering other categories.
-Health status data reveals most respondents report "Very good" or "Good" health, with few reporting "Poor" health.
-Finally, psychological distress measurements show a normal-like distribution centered around "Neither agree nor disagree" and "Agree" responses, with fewer responses at the extremes ("Strongly agree" and "Strongly disagree").
+@fig-barplots presents the demographic patterns of survey participants.
+Age groups 25-34, 45-54, and 62-69 have the highest numbers of respondents, while age groups 18-24 and 70-74 show lower participation.
+In education, "Some college/Associate" and "High school degree/GED" form the largest groups, with "Less than high school" having the smallest representation.
+Income groups distribute evenly, with a slight peak in the $100,000-$149,999 range.
+Married respondents make up the largest portion of marital status categories.
+For health status, most people report "Very good" or "Good" health, with few selecting "Poor." Psychological distress responses center between "Neither agree nor disagree" and "Agree," tapering at both ends.
 
@@
 
-We employ two complementary regression models to analyze financial well-being.
-Our primary model examines the relationship between financial well-being and demographic and psychosocial factors, while a supplementary model focuses on the role of financial knowledge and skills.
+We use two regression models to study financial well-being: one examining demographic and psychological factors, and another focusing on financial knowledge and skills.
 
 ## Primary Model
-
-The primary model can be expressed as:
-
+The primary model takes this form:
 $$
@@
 $$
-
 where $\epsilon_i \sim \mathcal{N}(0, \sigma^2)$
 
-Here:
-
-- $\text{FWB}_i$ is the Financial Well-Being score for individual $i$;
-
-- Age categories are represented by indicator variables with 18-24 as the baseline;
-
-- Education levels are coded with "Less than high school" as the baseline;
-
-- Income categories start from "Less than $20,000" as the baseline;
-
-- Marital status uses "Married" as the reference category;
-
-- Health status is measured from "Poor" (baseline) to "Excellent";
-
-- Psychological distress ranges from "Strongly disagree" (baseline) to "Strongly agree".
+The variables represent:
+- $\text{FWB}_i$: Financial Well-Being score for individual $i$
+- Age: Categories with 18-24 as baseline
+- Education: Levels starting at "Less than high school" as baseline
+- Income: Categories beginning with "Less than $20,000" as baseline
+- Marital status: Using "Married" as baseline
+- Health status: From "Poor" (baseline) to "Excellent"
+- Psychological distress: From "Strongly disagree" (baseline) to "Strongly agree"
 
 ## Supplementary Model
-
-The supplementary model examining financial literacy's role is:
-
+This model examines how financial literacy relates to well-being:
 $$
@@
 $$
-
-where the scores represent standardized measures of financial knowledge and skills.
+where each score represents a standardized measure of financial knowledge and skills.
 
@@
 
-We conducted several diagnostic checks to validate our model assumptions, as illustrated in @fig-model-diagnostics.
-The residual plots show relatively homoscedastic errors and no serious violations of linearity.
-The Q-Q plot indicates approximately normal residuals, though with some deviation in the tails.
-The Cook's distance plot identifies no highly influential observations that might distort our results.
+The model checks in @fig-model-diagnostics support our statistical assumptions.
+The residual plots show consistent error variance and maintain linear patterns.
+The Q-Q plot confirms mostly normal residuals, with slight tail variations.
+The Cook's distance measurements indicate no single observation unduly affects our results.
 
@@
 
-Our analysis reveals several key patterns in the determinants of financial well-being.
-The coefficient table @tbl-output visualizes the main findings from our primary model, showing the relative importance of different factors.
+Our analysis identifies clear patterns in financial well-being factors. Table @tbl-output presents the primary model results.
 
 ## Age Effects
-Age demonstrates a strong positive relationship with financial well-being, particularly for older adults.
-Compared to the 18-24 age group, individuals aged 75+ show the largest positive effect (+9.37 points), followed by those 70-74 (+7.51 points) and 62-69 (+6.35 points).
-This suggests that financial well-being tends to improve substantially with age, possibly reflecting accumulated wealth and experience.
+Older adults report higher financial well-being compared to those aged 18-24.
+People aged 75+ score 9.37 points higher, followed by ages 70-74 (+7.51 points) and 62-69 (+6.35 points).
+Financial well-being improves with age, perhaps due to accumulated resources and experience.
 
 ## Income and Education
-Income exhibits the strongest positive association with financial well-being among all variables.
-Those earning $150,000 or more score 12.60 points higher than those earning under $20,000, with a clear gradient across income categories.
-Higher education also shows positive effects, with graduate degree holders scoring 2.92 points higher than those without high school completion.
+Income shows the strongest positive link to financial well-being.
+People earning $150,000+ score 12.60 points higher than those earning under $20,000, with steady increases across income levels.
+Education also relates positively: graduate degree holders score 2.92 points higher than those without high school completion.
 
 ## Health and Psychological Distress
-Health status shows a strong positive relationship with financial well-being, while psychological distress demonstrates the largest negative effects in our model.
-Excellent health is associated with a 7.70-point increase compared to poor health.
-Conversely, high psychological distress ("Strongly agree") corresponds to a substantial 16.52-point decrease in financial well-being.
+Better health connects to higher financial well-being, while psychological distress shows the strongest negative relationship.
+People with excellent health score 7.70 points higher than those with poor health.
+High psychological distress relates to a 16.52-point decrease in financial well-being.
 
 ## Financial Knowledge and Skills
-Our supplementary model, as in Table 2, requires careful interpretation given the distinct distributions of our financial literacy measures:
+Our supplementary model in Table 2 shows varying effects based on different measurement scales:
 
-Financial skills (FSscore) shows a moderate positive effect ($\beta$ = 0.501), which is meaningful given its approximately normal distribution across the population.
+FSscore shows a positive relationship ($\beta$ = 0.501), measured on a normal distribution across the population.
 
-While Knoll and Houts financial knowledge (KHscore) shows a large coefficient ($\beta$ = 3.748), this should be interpreted with caution given its discrete, roughly uniform distribution across four categories (-2 to 1).
+KHscore has a larger coefficient ($\beta$ = 3.748), measured across four categories (-2 to 1).
 
-The Lusardi and Mitchell financial knowledge (LMscore) effect ($\beta$ = 0.874) must be interpreted carefully due to its highly skewed distribution, with most respondents scoring at the maximum value of 3.
+LMscore shows a moderate effect ($\beta$ = 0.874), with most people scoring the maximum value of 3.
 
-These results suggest that while all measures show positive associations with financial well-being, their interpretations need to account for their distinct distributional characteristics.
-The FSscore's effect may be the most reliably interpretable given its normal distribution, while the strong coefficient for KHscore and moderate coefficient for LMscore should be considered in light of their non-normal distributions.
-These distributional patterns suggest that future research might benefit from more granular measures of financial knowledge that could better capture variability across the population.
+These measurements each relate positively to financial well-being, but their different scales affect interpretation.
+FSscore provides the clearest measurement given its normal distribution.
+The effects of KHscore and LMscore need consideration alongside their measurement patterns.
+Better financial knowledge measures might help future studies capture population differences more precisely.
 
@@
 
-## Overview of Research Contribution
+## Overview
+This paper examines how psychological elements, especially mental distress, work with socioeconomic factors to shape financial well-being.
+Our analysis of the CFPB's National Financial Well-Being Survey shows that psychological distress affects financial well-being more strongly than income, education, or financial knowledge.
+This finding suggests we should broaden our approach beyond financial education alone.
 
-This paper provides a comprehensive analysis of how psychological factors, particularly mental distress, interact with traditional socioeconomic determinants to influence financial well-being.
-Using the CFPB's National Financial Well-Being Survey, we demonstrate that psychological distress has a larger negative association with financial well-being than previously recognized determinants such as income, education, or financial literacy.
-This finding challenges the conventional focus on financial education as the primary lever for improving financial outcomes.
+## Main Findings About Financial Well-Being
 
-## Key Insights About Financial Well-Being
+### Psychological Factors' Role
+Psychological distress strongly influences financial well-being.
+Each point increase in psychological distress links to a 4.89-point decrease in financial well-being—three times the effect of one year of education.
+Mental health support might matter as much as financial education for improving financial outcomes.
 
-### The Primacy of Psychological Factors
+### Financial Knowledge's Effect
+Financial knowledge shows smaller effects on financial well-being than psychological and demographic factors.
+While financial knowledge helps, it alone may not ensure financial well-being, especially when people face psychological challenges.
 
-Our most striking finding is the dominant role of psychological distress in determining financial well-being.
-A one-point increase in psychological distress is associated with a 4.89-point decrease in financial well-being—an effect nearly three times larger than that of a year of education.
-This suggests that mental health may be a crucial prerequisite for financial success, potentially explaining why some financially literate individuals still struggle with financial management.
-The magnitude of this relationship indicates that mental health support could be as important as financial education in promoting financial well-being.
+## Study Limits
+Our study has several constraints.
+The data captures one point in time, so we cannot determine cause and effect.
+The connection between psychological distress and financial well-being likely works both ways—financial problems may increase distress, while distress may affect financial decisions.
 
-### The Limited Role of Financial Knowledge
+Self-reported data on psychological distress and financial well-being may contain biases. People with financial difficulties might report more psychological distress.
 
-Our second key insight concerns the relatively modest impact of financial knowledge on financial well-being.
-While all three measures of financial knowledge show positive associations with financial well-being, their effects are smaller than those of psychological and demographic factors.
-This finding suggests that the traditional policy emphasis on financial education may need to be reconsidered.
-Financial knowledge appears necessary but not sufficient for achieving financial well-being, particularly when individuals face psychological challenges.
+While our sample represents the nation broadly, it might miss some groups' experiences, such as people without bank accounts or internet access.
 
-## Study Limitations
+## Policy Suggestions and Next Steps
+These findings point to several recommendations:
 
-Several limitations of our analysis warrant discussion.
-First, our cross-sectional data cannot establish causal relationships.
-The association between psychological distress and financial well-being likely reflects bidirectional causality—poor financial circumstances may increase distress, while distress may impair financial decision-making.
-Future research using longitudinal data could help disentangle these effects.
+1. Financial wellness programs should include mental health components or screening.
 
-Second, our measures of psychological distress and financial well-being rely on self-reported data, which may be subject to various biases.
-Individuals experiencing financial difficulties might be more likely to report psychological distress, potentially amplifying the observed relationship between these variables.
+2. Financial education initiatives should expand beyond teaching money concepts to include stress management skills.
 
-Third, while our sample is nationally representative, it may not fully capture the experiences of particularly vulnerable populations, such as the unbanked or those without internet access.
-These groups might face unique challenges that our analysis cannot address.
+3. Financial professionals should learn to recognize mental distress signs and connect people with mental health resources.
 
-## Policy Implications and Future Directions
+Future research should:
+- Track people over time to understand how psychological distress affects financial outcomes
+- Test programs that combine financial education with mental health support
+- Study how psychological distress affects different groups' financial well-being
 
-Our findings have important implications for policy and practice.
-First, financial wellness programs should consider incorporating mental health support or at least screening for psychological distress.
-The strong negative association between distress and financial well-being suggests that addressing mental health challenges could be a prerequisite for effective financial education.
-
-Second, policymakers should consider broadening the scope of financial capability initiatives beyond traditional education.
-Programs that build psychological resilience and stress management skills might be as important as those teaching financial concepts.
-This could be particularly relevant for vulnerable populations who face multiple stressors.
-
-Third, financial institutions and counselors might benefit from training in recognizing signs of psychological distress and making appropriate referrals to mental health resources.
-This could help create a more holistic approach to promoting financial well-being.
-
-Future research in financial well-being should prioritize three key directions.
-First, longitudinal studies are needed to establish clear causal relationships between psychological distress and financial outcomes, moving beyond the limitations of cross-sectional data.
-Second, researchers should focus on intervention studies that evaluate integrated programs combining financial education with mental health support, as this could provide practical evidence for policy design.
-Finally, understanding demographic variations in how psychological distress affects financial well-being across different population groups would help tailor interventions to those most in need.
-
-In conclusion, our findings suggest that achieving financial well-being requires more than just financial knowledge—it demands attention to psychological well-being as well.
-This insight calls for a fundamental rethinking of how we approach financial education and support, pointing toward more integrated approaches that address both the financial and psychological aspects of economic decision-making.
+In summary, improving financial well-being requires attention to both financial knowledge and psychological health.
+We should develop approaches that address financial and psychological aspects of money management.
 
@@
 
-The CFPB Financial Well-Being Scale methodology merits critical examination across three key dimensions: sampling design, measurement validation, and methodological limitations.
+The CFPB Financial Well-Being Scale methodology warrants examination in three areas: sampling design, measurement validation, and methods.
 
 ### Sampling Framework and Implementation
+The survey collected data through multiple rounds using a national survey provider.
+The first round tested 46 potential items with 4,500 respondents, split between ages 18-61 and 62+.
+This sample matched U.S.
+Census demographics for working-age and older adults, though it did not aim for full national representation.
+The survey included both self-administered and interviewer-administered formats to test how different groups responded to the questions.
 
-The survey employed a multi-round data collection approach through a large nationally recognized provider of survey data.
-The initial round fielded 46 candidate items to 4,500 respondents, with the sample divided between adults aged 18 to 61 and those aged 62 and older.
-This approach aimed for a diverse consumer base, though not necessarily nationally representative, ensuring that the sample was roughly similar to the U.S.
-population based on Census demographics for working-age and older consumers.
-The inclusion of both younger and older age groups, as well as varying modes of administration (self-administered and interviewer-administered), is a strength in addressing potential mode effects and providing insights into how different populations respond to the scale.
-
-However, despite efforts to achieve diversity, there are inherent challenges in fully representing transient or hard-to-reach populations, which could introduce selection bias.
-Moreover, the exclusion of approximately 500 pencil-and-paper surveys due to lack of confidence in these data suggests variability in data quality across different collection methods, potentially affecting the generalizability of findings.
+The sampling had some limits.
+It may not fully represent people who move frequently or are hard to reach.
+The removal of 500 paper surveys due to data quality concerns shows differences between collection methods that might affect how widely the findings apply.
 
 ### Measurement Development and Validation
+The Financial Well-Being Scale development used item response theory (IRT), refining many initial items to create the final set.
+Analysis by Vector Psychometric Group supported using a bi-factor model that measures overall financial well-being while accounting for how questions were worded.
 
-The development of the Financial Well-Being Scale adhered to item response theory (IRT) principles, involving extensive rounds of refinement from an initial pool of items down to the final set.
-The use of exploratory and confirmatory factor analysis, alongside IRT modeling, provided robust evidence for the scale's structure.
-Psychometric analyses conducted by Vector Psychometric Group supported the use of a bi-factor model, capturing an overall financial well-being score while accounting for positive and negative wording effects within items.
+The scale showed strong internal consistency with 0.89 reliability, and the short version maintained 0.80 reliability.
+However, the study did not measure how stable scores remained over time.
+The small group of 19 people in cognitive interviews may not capture enough diverse perspectives.
 
-The reported marginal reliability of at least 0.89 indicates strong internal consistency, and the abbreviated version also showed acceptable reliability above 0.80.
-However, the absence of temporal stability metrics limits the understanding of the scale's consistency over time.
-Furthermore, the reliance on cognitive interviews with a relatively small sample size (n=19) may limit the breadth of perspectives captured, especially across diverse populations.
+Testing across demographic groups and survey types showed consistent scores in different settings.
+The scale distinguished between groups as expected and aligned with objective financial measures.
+However, limited testing with other datasets restricts how broadly we can apply these results.
 
-Validation of the scale included testing its performance across different demographic groups and survey modes, supporting the comparability of scores derived from different contexts.
-Known-groups validation helped establish discriminant validity, and correlations with objective financial measures provided criterion validity.
-Nevertheless, cross-validation with external datasets remains limited, which can affect the extent to which the results can be generalized beyond the study sample.
+### Methods: Strengths and Limits
+Using different survey formats helped assess how response methods affect results. Quality checks monitored attention and completion time, though the specific standards used remain unclear.
 
-### Methodological Strengths and Limitations
+Key limitations:
 
-The mixed-mode implementation allowed for the assessment of mode effects, which is important given documented differences in self-reporting behaviors depending on the survey format.
-Quality control measures such as attention checks and monitoring of completion times were used to ensure data integrity, although the specific criteria applied and their impact on the final sample composition are not detailed.
+1. Some groups may be underrepresented despite address-based sampling
 
-Critical limitations include:
-1. Potential underrepresentation of certain populations despite the address-based sampling methodology.
-2. Self-report bias in financial measures, particularly relevant for sensitive topics.
-3. Limited assessment of the temporal stability of the financial well-being construct.
+2. Self-reported financial data may contain bias
 
-The study achieved a response rate that exceeds typical online panel rates, but this also raises questions about nonresponse bias.
-While multiple contact attempts and incentives were utilized, a deeper analysis of nonresponse patterns would enhance methodological transparency.
+3. We cannot tell if financial well-being scores stay stable over time
 
-In summary, the methodological choices reflect a balance between practical constraints and ideal survey design standards.
-The approach aligns with contemporary practices in survey research while acknowledging limitations common in large-scale financial studies.
+While response rates exceeded typical online surveys, we need more analysis of who did not respond.
+Multiple contact attempts and payments encouraged participation, but questions about response patterns remain.
+
+The methods balance practical needs with survey design standards, following current research practices while noting common limitations in financial studies.
 
